---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# shinymcp

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/shinymcp)](https://CRAN.R-project.org/package=shinymcp)
[![R-CMD-check](https://github.com/JamesHWade/shinymcp/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JamesHWade/shinymcp/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/JamesHWade/shinymcp/graph/badge.svg)](https://app.codecov.io/gh/JamesHWade/shinymcp)
<!-- badges: end -->

shinymcp converts [Shiny](https://shiny.posit.co/) apps into
[MCP Apps](https://modelcontextprotocol.io/) --- interactive UIs that render
directly inside AI chat interfaces like Claude Desktop.

## Installation

You can install the development version of shinymcp from
[GitHub](https://github.com/) with:

```r
# install.packages("pak")
pak::pak("JamesHWade/shinymcp")
```

## Quick start

An MCP App has two parts: **UI components** that render in the chat interface,
and **tools** that run R code when inputs change.

```{r example-scratch, eval = FALSE}
library(shinymcp)
library(bslib)

ui <- page_sidebar(
  theme = bs_theme(preset = "shiny"),
  title = "Dataset Explorer",
  sidebar = sidebar(
    mcp_select("dataset", "Choose dataset", c("mtcars", "iris", "pressure"))
  ),
  card(
    card_header("Summary"),
    mcp_text("summary")
  )
)

tools <- list(
  ellmer::tool(
    fun = function(dataset = "mtcars") {
      data <- get(dataset, envir = asNamespace("datasets"))
      paste(capture.output(summary(data)), collapse = "\n")
    },
    name = "get_summary",
    description = "Get summary statistics for the selected dataset",
    arguments = list(
      dataset = ellmer::type_string("Dataset name")
    )
  )
)

app <- mcp_app(ui, tools, name = "dataset-explorer")
serve(app)
```

Save this as `app.R`, then register it in your Claude Desktop config:

```json
{
  "mcpServers": {
    "dataset-explorer": {
      "command": "Rscript",
      "args": ["/path/to/app.R"]
    }
  }
}
```

Restart Claude Desktop. When the tool is invoked, an interactive UI appears
inline in the conversation. Changing the dropdown calls the tool and updates
the output --- no page reload needed.

## Converting a Shiny app

The core idea: **flatten your reactive graph into tool functions**.

Each connected group of inputs → reactives → outputs becomes a single tool
that takes input values as arguments and returns a named list of outputs.

```{r shiny-to-mcp, eval = FALSE}
# --- Shiny ---
server <- function(input, output, session) {
  filtered <- reactive({
    penguins[penguins$species == input$species, ]
  })
  output$scatter <- renderPlot({
    ggplot(filtered(), aes(x, y)) + geom_point()
  })
  output$stats <- renderPrint({
    summary(filtered())
  })
}

# --- MCP App tool ---
ellmer::tool(
  fun = function(species = "Adelie") {
    filtered <- penguins[penguins$species == species, ]

    # Render plot to base64 PNG
    tmp <- tempfile(fileext = ".png")
    ggplot2::ggsave(tmp, my_plot, width = 7, height = 4, dpi = 144)
    on.exit(unlink(tmp))

    list(
      scatter = base64enc::base64encode(tmp),
      stats = paste(capture.output(summary(filtered)), collapse = "\n")
    )
  },
  name = "explore",
  description = "Filter and visualize penguins",
  arguments = list(
    species = ellmer::type_string("Penguin species")
  )
)
```

Return keys (`scatter`, `stats`) must match output IDs in the UI
(`mcp_plot("scatter")`, `mcp_text("stats")`). The bridge routes each
value to the correct output element.

For a full worked example converting a Shiny app step-by-step, see
`vignette("converting-shiny-apps")`.

## Component reference

shinymcp provides input and output components that mirror Shiny functions:

### Inputs

| Shiny | shinymcp |
|---|---|
| `selectInput()` | `mcp_select()` |
| `textInput()` | `mcp_text_input()` |
| `numericInput()` | `mcp_numeric_input()` |
| `checkboxInput()` | `mcp_checkbox()` |
| `sliderInput()` | `mcp_slider()` |
| `radioButtons()` | `mcp_radio()` |
| `actionButton()` | `mcp_action_button()` |

### Outputs

| Shiny | shinymcp | Notes |
|---|---|---|
| `textOutput()` / `verbatimTextOutput()` | `mcp_text()` | Renders in `<pre>` with monospace font |
| `plotOutput()` | `mcp_plot()` | Tool returns base64-encoded PNG |
| `tableOutput()` | `mcp_table()` | Tool returns HTML table string |
| `htmlOutput()` | `mcp_html()` | Tool returns raw HTML |

Each component generates HTML with `data-shinymcp-*` attributes that the
JS bridge reads at runtime:

```{r components, eval = FALSE}
mcp_select("color", "Color", c("red", "green", "blue"))
# <select data-shinymcp-input="color" ...>

mcp_plot("my_plot")
# <div data-shinymcp-output="my_plot" data-shinymcp-output-type="plot" ...>
```

## Examples

shinymcp ships with example apps:
```{r examples, eval = FALSE}
# Minimal: select input + text output
system.file("examples", "hello-mcp", "app.R", package = "shinymcp")

# Full dashboard: Palmer Penguins with ggplot2 scatter plot, multiple
# inputs (species filter, axis selectors, trend line toggle), and
# summary statistics
system.file("examples", "penguins", "app.R", package = "shinymcp")
```

## How it works

MCP Apps render inside sandboxed iframes in the AI chat interface.
A lightweight JavaScript bridge (no npm dependencies) handles
communication via postMessage/JSON-RPC:

1. User changes an input → bridge collects all input values
2. Bridge sends a `tools/call` request to the host
3. Host proxies the call to the MCP server (your R process)
4. R tool function runs, returns results
5. Bridge updates output elements with the response

The bridge also implements the MCP Apps initialization handshake
(`ui/initialize`), auto-resize notifications, and teardown handling.

## Related packages

- [ellmer](https://ellmer.tidyverse.org/) --- LLM framework that shinymcp uses
  for tool definitions
- [bslib](https://rstudio.github.io/bslib/) --- Bootstrap-based layout and
  theming for the UI
- [mcptools](https://github.com/posit-dev/mcptools) --- MCP server framework
  (shinymcp extends this with resource support)
- [deputy](https://github.com/JamesHWade/deputy) --- Agentic AI framework
  with a bundled skill for AI-assisted Shiny conversion
