---
title: "Adding your app to Claude Desktop, VS Code, or Goose"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding your app to Claude Desktop, VS Code, or Goose}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Up to now you have previewed your app with `preview_app()`. That opens a
browser window that stands in for an AI assistant, so you can click the inputs
and watch the card update. It is a good rehearsal. It is still a rehearsal.

The reason MCP exists is the next step: a real assistant, running on your
machine, calls your R code and shows the result in the conversation. No browser
tab of your own, no server you babysit. You write the tool, register it once,
and the assistant does the rest.

This chapter connects one app to three clients: Claude Desktop, VS Code, and
Goose. The app is the same in every case. What changes is a few lines in a
config file. Once you have done it once, the other two take a minute each.

## What a client actually talks to

An MCP client does not load your package, and it does not call a web service you
host. It launches a command and talks to that command's standard input and
output. The messages are [JSON-RPC](https://www.jsonrpc.org/specification): after
an `initialize` handshake, the client sends `tools/list` to learn what your app
offers and `tools/call` to run a tool. A client that supports MCP UI also reads
the app's `ui://` resource to get the card's HTML. Your job is to provide a
command that speaks that protocol.

In shinymcp the command is `Rscript` and the script is a few lines that build an
app and call `serve()`. That is the whole server. There is no daemon to keep
alive and no port to remember, because the client starts the process when it
needs your tools and stops it when it is done.

Two facts follow from this, and they explain most of what can go wrong later:

1.  The client runs `Rscript`, so `Rscript` has to be on a `PATH` the client can
    see, and the R library it uses has to contain shinymcp and ellmer.
2.  The client passes your script as a file path, so that path has to be
    absolute. A relative path is meaningless to a program the client launched
    from somewhere else.

## The server you will connect

shinymcp ships a small example built for exactly this. Find it on disk with:

```{r}
system.file("examples", "serve-to-client", "serve.R", package = "shinymcp")
```

Open that file and you will see an ordinary `mcp_app()` with one tool, ending in
a single new line:

```{r}
library(shinymcp)

app <- mcp_app(
  ui = htmltools::tagList(
    mcp_text_input("name", "Your name", value = "world"),
    mcp_text("greeting")
  ),
  tools = list(
    ellmer::tool(
      fun = function(name = "world") {
        list(greeting = paste0("Hello, ", name, "!"))
      },
      name = "greet",
      description = "Greet a person by name",
      arguments = list(
        name = ellmer::type_string("The name to greet")
      )
    )
  ),
  name = "shinymcp-demo"
)

serve(app, type = "stdio")
```

The only line that is new is the last one. `serve(app, type = "stdio")` reads
JSON-RPC from standard input and writes responses to standard output, which is
the channel every client in this chapter uses. (`serve()` can also run over
HTTP; that is for a later section.)

Before you touch any client, confirm the script runs:

```bash
which Rscript
Rscript "$(Rscript -e 'cat(system.file("examples","serve-to-client","serve.R",package="shinymcp"))')"
```

The second command will print a line about serving over stdio and then sit
there. That is correct: it is waiting for a client to speak to it on stdin.
Press Ctrl-C to get your prompt back. If instead you get an error, fix that now.
A client cannot do anything with a server that will not start.

That snippet is written for macOS and Linux shells. On Windows, use
`where.exe Rscript` (or `Get-Command Rscript` in PowerShell) to find Rscript, and
run the `system.file()` call in R to get the path to `serve.R`.

You will reuse one thing in all three clients: the absolute path to `serve.R`.
Copy the output of the `system.file()` call above and keep it handy. The
examples below write it as `/ABSOLUTE/PATH/TO/serve.R`.

## Claude Desktop

Claude Desktop keeps its MCP servers in a JSON file. You can edit it through the
app or open it directly.

1.  In the app, go to **Settings, Developer, Edit Config**. Or open the file
    yourself:

    -   macOS: `~/Library/Application Support/Claude/claude_desktop_config.json`
    -   Windows: `%APPDATA%\Claude\claude_desktop_config.json`

2.  Add your server under `mcpServers`:

    ```json
    {
      "mcpServers": {
        "shinymcp-demo": {
          "command": "Rscript",
          "args": ["/ABSOLUTE/PATH/TO/serve.R"]
        }
      }
    }
    ```

    If the file already has other servers, add `shinymcp-demo` alongside them
    rather than replacing the block.

3.  Quit Claude Desktop and open it again. Config changes are read at startup, so
    a reload is not optional.

Now ask Claude to greet someone. It finds the `greet` tool, calls it, and the
card renders inline in the conversation. Type a name into the card and the
greeting updates, because each edit is another `tools/call` to your R function.

> If `command` is bare `Rscript` and nothing happens, Claude is probably
> launched with a `PATH` that does not include your R. Replace `"Rscript"` with
> the absolute path, for example `"/opt/homebrew/bin/Rscript"` on Apple silicon.
> `which Rscript` prints the path to use (`where.exe Rscript` on Windows).

## VS Code

VS Code can read MCP servers from a per-workspace file, which is handy because
the server then travels with the project you put it in.

1.  Create `.vscode/mcp.json` in your workspace:

    ```json
    {
      "servers": {
        "shinymcp-demo": {
          "type": "stdio",
          "command": "Rscript",
          "args": ["/ABSOLUTE/PATH/TO/serve.R"]
        }
      }
    }
    ```

    Two things differ from Claude Desktop: the key is `servers`, not
    `mcpServers`, and each server names its `type` explicitly.

2.  Open `.vscode/mcp.json` and click **Start** on the server, or run **MCP:
    List Servers** from the Command Palette and start it from there.

3.  Open Copilot Chat and switch it to **Agent** mode. The `greet` tool is now in
    the tool list, and the agent can call it.

VS Code calls the tool and shows you its result. Whether it draws the
interactive card depends on its MCP UI support; the tool itself works either way.

## Goose

[Goose](https://block.github.io/goose/) calls MCP servers "extensions." You can
add one with the `goose configure` wizard or by editing the config file. The
wizard writes the same file, so it pays to read it once.

Goose's config lives at:

-   macOS and Linux: `~/.config/goose/config.yaml`
-   Windows: `%APPDATA%\Block\goose\config\config.yaml`

Add your server under `extensions`:

```yaml
extensions:
  shinymcp-demo:
    type: stdio
    enabled: true
    name: shinymcp-demo
    cmd: Rscript
    args:
      - /ABSOLUTE/PATH/TO/serve.R
    timeout: 300
```

A few field names are particular to Goose. The command key is `cmd`, not
`command`. `enabled: true` is what actually turns the extension on. `timeout` is
in seconds and gives the R process room to start and answer.

If you prefer the wizard, run `goose configure`, choose **Add Extension**, then
**Command-line Extension**, and give it `Rscript` as the command with the
absolute path to `serve.R` as the argument. It writes the same block for you.

Start a Goose session and the `greet` tool is available. As with VS Code, Goose
runs the tool and shows the result; the card renders where Goose supports MCP
UI.

## Will the card actually render?

The answer is not the same for every client.

Your app has two layers. The lower layer is the tool: a function that takes
arguments and returns a named list. Every MCP client in this chapter can list
and call it, and that contract is stable across all of them. The upper layer is
the card: the HTML interface served as a `ui://` resource, with the JavaScript
bridge that reads inputs and patches outputs. A client only draws that card if
it implements the MCP UI part of the spec.

Today Claude Desktop renders the card inline. VS Code and Goose reliably call
the tool and show its result, and their UI support is improving. So when you
test a new client and see the greeting as text rather than a live card, the
connection is working. You are looking at the lower layer doing its job while the
upper layer waits for the host to catch up.

So write your tool so its return value reads well as plain text. The card is a
better experience where you can get it, and the plain result works everywhere
else.

## When stdio is not enough

Everything above uses `type = "stdio"`, where the client owns the R process and
each client gets its own. That is the right default for a tool you run on your
own machine.

`serve()` also speaks HTTP:

```{r}
serve(app, type = "http", port = 8080)
```

This starts a long-running server that accepts JSON-RPC over POST on `127.0.0.1`.
That suits a shared or remote deployment where you would rather run one R process
than have every client spawn its own, though reaching it from another machine
means putting a proxy in front. It is a different operational shape: you manage
the process, the port, and the network, and clients connect to an MCP endpoint
rather than launching a command. Reach for it when you have outgrown a local
script, not before.

## Troubleshooting

When a client shows nothing, work from the process outward.

**The server will not start on its own.** Run `Rscript /path/to/serve.R` in a
terminal. If that errors, no client can help. The usual cause is a missing
package: install shinymcp and ellmer into the library that this `Rscript` uses,
which you can check with `Rscript -e '.libPaths()'`.

**"command not found" or silence.** The client's `PATH` differs from your
shell's. Use the absolute path to `Rscript` (from `which Rscript`, or
`where.exe Rscript` on Windows) as the command in the config.

**The tool runs but errors.** Read the client's MCP logs. Your server writes
diagnostics to standard error, and the client surfaces them there: VS Code under
**MCP: List Servers** then the server's output, Claude Desktop in its logs
folder, Goose in its session output.

**Edits to the config do nothing.** Most clients read their config at startup.
Restart Claude Desktop, start the VS Code server again, and start a fresh Goose
session.

## Where to go next

You now have a working loop: write a tool, point a client at it, call it from a
conversation. The other vignettes fill in what goes inside that loop.
[shinymcp by example](shinymcp-by-example.html) walks through the components and
patterns one at a time, and [Converting Shiny Apps to MCP
Apps](converting-shiny-apps.html) shows how to turn an app you already have into
one of these.
