---
title: "Compelling shinymcp use cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compelling shinymcp use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`shinymcp` is strongest when an AI assistant needs to do more than print a
number. The assistant can call a tool, receive structured data for reasoning,
and show the user an interactive card with controls, tables, plots, and
formatted output.

The package includes a use-case gallery with three small but realistic MCP
Apps:

| Use case | What it demonstrates |
|---|---|
| Revenue scenario board | Scenario controls, model-facing structured values, a table, and an ARR plot |
| Experiment planner | Statistical planning, a power curve, and a concise recommendation |
| Incident triage console | Operational decision support, status HTML, and a runbook table |

## Run a use case as an MCP App

The gallery entrypoint serves one use case at a time. Set
`SHINYMCP_USE_CASE` to `revenue`, `experiment`, or `incident`.

```{bash}
SHINYMCP_USE_CASE=revenue Rscript \
  "$(Rscript -e 'cat(system.file("examples", "use-cases", "app.R", package = "shinymcp"))')"
```

For Claude Desktop or another MCP host, point the server command at that same
`app.R` file and set the environment variable for the use case you want to
show.

## Revenue scenario board

This card turns go-to-market assumptions into a twelve-month forecast. The
assistant can reason over the structured result while the user sees the ARR
trajectory and monthly forecast table.

```{r}
source(system.file("examples", "use-cases", "apps.R", package = "shinymcp"))
app <- shinymcp_use_case("revenue")
app$call_tool("forecast_revenue", list(
  segment = "Mid-market",
  visitors = 25000,
  trial_rate = 7,
  win_rate = 22,
  contract_value = 9000,
  monthly_churn = 2.5
))
```

## Experiment planner

This card helps an assistant move from "can we detect a 15% lift?" to a
specific sample size and runtime. The plot gives the human reviewer a quick
sense of how much margin the design has.

```{r}
app <- shinymcp_use_case("experiment")
app$call_tool("plan_experiment", list(
  baseline_rate = 12,
  minimum_effect = 15,
  target_power = 0.9,
  traffic_per_day = 6000
))
```

## Incident triage console

This card is useful for support, SRE, and operations chats. The assistant can
collect facts conversationally, call the tool, and return a priority, briefing,
and runbook without losing the structured values it needs for follow-up.

```{r}
app <- shinymcp_use_case("incident")
app$call_tool("triage_incident", list(
  service = "Payments",
  severity = "Degraded",
  affected_users = 1200,
  minutes_open = 24,
  regulated_data = TRUE
))
```

## shinychat integration

The gallery also includes a Shiny app that registers all three cards as
`shinychat` tool results through `as_shinychat_tool()`.

```{bash}
Rscript "$(Rscript -e 'cat(system.file("examples", "use-cases", "shinychat-app.R", package = "shinymcp"))')"
```

That app uses `chat_mod_ui()` and `chat_mod_server()` with one ellmer client per
Shiny session. It expects a configured provider for
`ellmer::chat("openai/gpt-4.1-nano")`; change the model string if your local
demo uses another provider.

## Why these examples matter

These examples cover the surfaces most teams need before adopting MCP Apps:

- Rich UI for people: inputs, tables, plots, and formatted HTML.
- Structured values for the model: the assistant can reason over numbers
  without scraping the display.
- Portable deployment: the same `McpApp` can be served to an MCP host, embedded
  in Shiny, or wrapped as a shinychat tool card.
- Small cards over giant dashboards: each card has one job and maps cleanly to
  one tool call.
