---
title: "Converting Shiny Apps to MCP Apps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting Shiny Apps to MCP Apps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

An MCP App is an interactive UI that renders inside an AI chat interface
like Claude Desktop. A Shiny app runs in a browser tab backed by a
persistent server. An MCP App is a small HTML document, and user
interactions trigger tool calls: stateless R functions that the AI host
invokes on demand.

This vignette converts a Shiny app into an MCP App with shinymcp, using
the classic Palmer Penguins explorer as the worked example.

## Key differences from Shiny

Before converting, it helps to understand what changes:

| | Shiny | MCP App |
|---|---|---|
| **Runs in** | Browser tab | AI chat interface (iframe) |
| **Server** | Persistent R process with reactive graph | Stateless tool calls |
| **Reactivity** | `reactive()`, `observe()`, `render*()` | Tool function called on input change |
| **UI framework** | shiny.js + WebSocket | Plain HTML + postMessage bridge |
| **State** | Server-side reactive values | Recomputed each tool call |
| **Layout** | `fluidPage()`, `sidebarLayout()` | bslib (`page_sidebar()`, `card()`) |

The work of converting an app is flattening your reactive graph into tool
functions. Each connected group of inputs → reactives → outputs becomes a
single tool that takes the inputs as arguments and returns the outputs as
a named list.

## Step 1: Identify inputs and outputs

Start with your Shiny app and list every input and output. Here's a
typical penguins Shiny app:

```{r}
# --- Original Shiny app ---
library(shiny)
library(ggplot2)
library(palmerpenguins)

ui <- fluidPage(
  titlePanel("Palmer Penguins Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("species", "Species", c("All", "Adelie", "Chinstrap", "Gentoo")),
      selectInput("x_var", "X axis", c("bill_length_mm", "bill_depth_mm",
                                        "flipper_length_mm", "body_mass_g")),
      selectInput("y_var", "Y axis", c("bill_depth_mm", "bill_length_mm",
                                        "flipper_length_mm", "body_mass_g")),
      checkboxInput("trend", "Show trend line", FALSE)
    ),
    mainPanel(
      plotOutput("scatter"),
      verbatimTextOutput("stats")
    )
  )
)

server <- function(input, output, session) {
  filtered_data <- reactive({
    data <- penguins[complete.cases(penguins), ]
    if (input$species != "All") {
      data <- data[data$species == input$species, ]
    }
    data
  })

  output$scatter <- renderPlot({
    p <- ggplot(filtered_data(),
                aes(x = .data[[input$x_var]], y = .data[[input$y_var]],
                    color = species)) +
      geom_point(alpha = 0.7, size = 2.5)
    if (input$trend) p <- p + geom_smooth(method = "lm", se = FALSE)
    p
  })

  output$stats <- renderPrint({
    summary(filtered_data()[, c(input$x_var, input$y_var, "species")])
  })
}

shinyApp(ui, server)
```

This app has four inputs: a `selectInput` for `species`, two more for
`x_var` and `y_var`, and a `checkboxInput` for `trend`. It has two
outputs: a `plotOutput` named `scatter` holding a ggplot, and a
`verbatimTextOutput` named `stats` holding summary statistics.

The reactive logic is one piece. `filtered_data()` filters penguins by
species, and both outputs depend on it together with the axis and trend
inputs. Everything is connected, so it maps to a single tool.

## Step 2: Map components

The JS bridge auto-detects inputs by matching tool argument names to element
`id` attributes. You can keep your Shiny inputs as they are, as long as the
`id` matches a tool argument name:

| Shiny input | What to do |
|---|---|
| `selectInput("species", "Species", choices)` | Keep it. Auto-detected by `id="species"` |
| `selectInput("x_var", "X axis", choices)` | Keep it. Auto-detected by `id="x_var"` |
| `checkboxInput("trend", "Show trend line")` | Keep it. Auto-detected by `id="trend"` |

For outputs, replace Shiny outputs with shinymcp equivalents:

| Shiny output | shinymcp |
|---|---|
| `plotOutput("scatter")` | `mcp_plot("scatter")` |
| `verbatimTextOutput("stats")` | `mcp_text("stats")` |

The IDs stay the same. If you need to mark a non-standard element as an
output, use `mcp_output(tag, id, type)`.

## Step 3: Build the UI

Use bslib for layout instead of Shiny's `fluidPage()` / `sidebarLayout()`.
bslib components work directly with htmltools, with no Shiny server. Standard
`shiny::selectInput()`, `shiny::checkboxInput()`, and the rest are
auto-detected by the bridge:

```{r}
library(shinymcp)
library(bslib)
library(htmltools)

ui <- page_sidebar(
  theme = bs_theme(preset = "shiny"),
  title = "Palmer Penguins Explorer",
  sidebar = sidebar(
    width = 260,
    shiny::selectInput(
      "species", "Species",
      c("All", "Adelie", "Chinstrap", "Gentoo")
    ),
    shiny::selectInput("x_var", "X axis", c(
      "Bill Length (mm)" = "bill_length_mm",
      "Bill Depth (mm)" = "bill_depth_mm",
      "Flipper Length (mm)" = "flipper_length_mm",
      "Body Mass (g)" = "body_mass_g"
    )),
    shiny::selectInput("y_var", "Y axis", c(
      "Bill Depth (mm)" = "bill_depth_mm",
      "Bill Length (mm)" = "bill_length_mm",
      "Flipper Length (mm)" = "flipper_length_mm",
      "Body Mass (g)" = "body_mass_g"
    )),
    shiny::checkboxInput("trend", "Show trend line")
  ),
  card(
    card_header("Scatter Plot"),
    mcp_plot("scatter", height = "380px")
  ),
  card(
    card_header("Summary Statistics"),
    mcp_text("stats")
  )
)
```

Named choice vectors (e.g. `"Bill Length (mm)" = "bill_length_mm"`)
work exactly like Shiny: the name is displayed, the value is sent
to the tool.

## Step 4: Convert reactive logic to a tool

Flatten the reactive graph into a single function. The arguments are the
input values, each with a sensible default. The body does what the
reactives and renderers did. The function returns a named list that maps
output IDs to values.

For plots, render to a temporary PNG and return the base64-encoded image.
The bridge displays it as an `<img>` element.

```{r}
tools <- list(
  ellmer::tool(
    fun = function(
      species = "All",
      x_var = "bill_length_mm",
      y_var = "bill_depth_mm",
      trend = FALSE
    ) {
      data <- palmerpenguins::penguins
      data <- data[complete.cases(data), ]

      # Filter (was: filtered_data reactive)
      if (species != "All") {
        data <- data[data$species == species, ]
      }

      # Plot (was: renderPlot)
      p <- ggplot2::ggplot(
        data,
        ggplot2::aes(x = .data[[x_var]], y = .data[[y_var]], color = species)
      ) +
        ggplot2::geom_point(alpha = 0.7, size = 2.5) +
        ggplot2::theme_minimal(base_size = 13)

      if (isTRUE(trend)) {
        p <- p + ggplot2::geom_smooth(method = "lm", se = FALSE)
      }

      # Render plot to base64 PNG
      tmp <- tempfile(fileext = ".png")
      ggplot2::ggsave(tmp, p, width = 7, height = 4, dpi = 144, bg = "white")
      on.exit(unlink(tmp))
      plot_b64 <- base64enc::base64encode(tmp)

      # Summary text (was: renderPrint)
      stats <- paste(capture.output({
        cat(sprintf("Observations: %d penguins\n\n", nrow(data)))
        print(summary(data[, c(x_var, y_var, "species")]))
      }), collapse = "\n")

      # Return named list: keys must match output IDs
      list(scatter = plot_b64, stats = stats)
    },
    name = "explore_penguins",
    description = "Explore the Palmer Penguins dataset with scatter plots",
    arguments = list(
      species = ellmer::type_string("Species filter: All, Adelie, Chinstrap, or Gentoo"),
      x_var = ellmer::type_string("X axis variable name"),
      y_var = ellmer::type_string("Y axis variable name"),
      trend = ellmer::type_boolean("Whether to show a linear trend line")
    )
  )
)
```

A few things are worth getting right here.

The return keys match the output IDs. `list(scatter = ..., stats = ...)`
corresponds to `mcp_plot("scatter")` and `mcp_text("stats")`, and the
bridge uses these keys to route values to the right output elements.

Plots travel as base64. Use `ggsave()` or `png()` plus `dev.off()`, then
`base64enc::base64encode()`, and the bridge wraps the result in an `<img>`
tag. Text output renders in a `<pre>` tag, so the column alignment from
R's `summary()` and `cat()` survives.

Every argument needs a default value. The tool uses the defaults when it
first loads, and they tell the AI what format to expect.

Tool calls are stateless: no `reactiveVal()`, no `<<-`, no session state.
Each call recomputes from scratch, which is fine for most apps because the
calls are fast.

## Step 5: Assemble and serve

Wire up the UI, tools, and server:

```{r}
app <- mcp_app(ui, tools, name = "penguins-explorer")
serve(app)
```

Save this as a single `app.R` file. To register it with Claude Desktop,
add an entry to your Claude Desktop config
(`~/Library/Application Support/Claude/claude_desktop_config.json` on macOS):

```json
{
  "mcpServers": {
    "penguins": {
      "command": "/opt/homebrew/bin/Rscript",
      "args": ["/path/to/your/app.R"]
    }
  }
}
```

Restart Claude Desktop, and the app will appear as an interactive UI
when the `explore_penguins` tool is called.

## Conversion patterns

### One reactive group → one tool

If all your outputs share the same reactive dependencies, they belong
in a single tool. This is the most common case:

```{r}
# Shiny: two outputs from one reactive
server <- function(input, output, session) {
  data <- reactive({ mtcars[mtcars$cyl == input$cyl, ] })
  output$plot <- renderPlot({ plot(data()$mpg) })
  output$text <- renderText({ paste(nrow(data()), "cars") })
}

# MCP App: one tool returning both
ellmer::tool(
  fun = function(cyl = 4) {
    data <- mtcars[mtcars$cyl == cyl, ]
    list(
      plot = render_plot_base64(function() plot(data$mpg)),
      text = paste(nrow(data), "cars")
    )
  },
  # ...
)
```

### Independent outputs → multiple tools

If outputs have completely independent inputs, use separate tools.
The bridge calls all tools whose inputs changed:

```{r}
# Tool 1: only uses the "dataset" input
ellmer::tool(
  fun = function(dataset = "mtcars") { ... },
  name = "summarize_data",
  # ...
)

# Tool 2: only uses the "n" input
ellmer::tool(
  fun = function(n = 100) { ... },
  name = "generate_sample",
  # ...
)
```

### Replacing `reactiveVal` / session state

MCP tool calls are stateless. If your Shiny app uses `reactiveVal()`
or `reactiveValues()` to accumulate state across interactions, you have
two options. Most filtering and selection state can be derived from the
current input values alone, so recompute it. For state you genuinely have
to carry, like a todo list, write it to a temp file and read it back on
the next tool call.

### Tables

For `tableOutput`, return an HTML table string. `knitr::kable()` is
the easiest approach:

```{r}
fun = function(dataset = "mtcars") {
  data <- head(get(dataset, envir = asNamespace("datasets")), 10)
  list(
    my_table = knitr::kable(data, format = "html")
  )
}
```

Use `mcp_table("my_table")` in the UI. The bridge sets `innerHTML`
directly.

### Dynamic UI

MCP Apps don't support `renderUI()` / `uiOutput()`. Instead:

- Use `mcp_html(id)` and return HTML strings from the tool
- For show/hide, return an empty string when the element should be hidden
- For dynamic choices, use a fixed `mcp_select()` and document
  valid values in the tool description

## Layout with bslib

Since MCP Apps use htmltools directly (not Shiny's `fluidPage`), use
bslib for layout. Common patterns:

```{r}
# Sidebar layout (standard shiny inputs auto-detected)
page_sidebar(
  sidebar = sidebar(
    shiny::selectInput("x", "Variable", choices),
    shiny::checkboxInput("log", "Log scale")
  ),
  card(mcp_plot("main_plot")),
  card(mcp_text("summary"))
)

# Multi-column layout
page(
  theme = bs_theme(preset = "shiny"),
  layout_columns(
    col_widths = c(6, 6),
    card(card_header("Plot"), mcp_plot("plot1")),
    card(card_header("Table"), mcp_table("table1"))
  )
)

# Stacked cards
page(
  card(card_header("Controls"), shiny::selectInput("x", "Pick", c("a", "b"))),
  card(card_header("Output"), mcp_text("result"))
)
```

All CSS and JS dependencies from bslib are automatically inlined into
the MCP App's HTML resource.

## What can't be converted

Some Shiny features don't have MCP App equivalents:

- **`fileInput`**: MCP Apps can't receive file uploads. Use
  `mcp_text_input()` for file paths instead, or have the tool read
  from a known directory.
- **Shiny modules with internal state**: Flatten the module logic into
  top-level tools.
- **`invalidateLater` / polling**: MCP tools are request-response only.
  No background updates.
- **JavaScript-heavy widgets** (DT, plotly, leaflet): These require
  Shiny's JS runtime. Use static alternatives (base R plots,
  `knitr::kable` tables).

## Full example

The complete converted penguins app is available at:

```{r, eval = FALSE}
system.file("examples", "penguins", "app.R", package = "shinymcp")
```
