---
title: "Running an app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running an app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

An MCP App is served by an MCP server. A chat client connects to it in one
of two ways: it starts the server itself and talks to it over standard input
and output (stdio), or it connects to a URL (Streamable HTTP). shinymcp
serves apps both ways, and inside a Shiny app for Posit Connect.

Every way below works with an [`McpApp`](../reference/McpApp.html), a Shiny
app object, a path to a Shiny app's directory, or a list of apps to serve
from one server.

## While you work: `preview_app()`

```{r}
preview_app(app)
```

`preview_app()` serves the app over HTTP on your machine and opens a page
that plays the part of a chat client: it calls the app's tool the way a
model would, shows the app in a sandboxed frame under the same Content
Security Policy clients use, and lists what the model received, each tool
call, and the protocol messages. Change the arguments and call again, or
switch the theme and width to see the app in a dark or narrow chat.

`preview_app()` returns right away. Call `$stop()` on its result when you're
done.

## Desktop clients, over stdio

Claude Desktop, Claude Code, VS Code, Goose, and most other desktop clients
start an MCP server as a program. Write the app in a script that ends with
`serve()`:

```{r}
# ~/apps/mileage/app.R
library(shiny)
library(shinymcp)

ui <- fluidPage(...)
server <- function(input, output, session) { ... }

serve(as_mcp_app(shinyApp(ui, server), name = "mileage"))
```

Run it in a terminal first: `Rscript ~/apps/mileage/app.R` should sit waiting
for input (Ctrl+C stops it). An error there is the error the client would
hit, without the client's log in the way.

The protocol runs over standard output, so the script must not print to it
before `serve()` starts; messages and warnings go to standard error, which
is fine. While a request is being handled, anything the app prints is sent
to standard error, where clients keep it in their logs.

Then register the script with the client, using full paths. On macOS,
desktop apps don't see your shell's `PATH`, so give the full path to
`Rscript` too (`which Rscript` prints it).

**Claude Desktop**, under Settings > Developer > Edit Config:

```json
{
  "mcpServers": {
    "mileage": {
      "command": "/usr/local/bin/Rscript",
      "args": ["/Users/you/apps/mileage/app.R"]
    }
  }
}
```

**Claude Code**:

```sh
claude mcp add mileage -- Rscript /Users/you/apps/mileage/app.R
```

**VS Code**, in `.vscode/mcp.json`:

```json
{
  "servers": {
    "mileage": {
      "type": "stdio",
      "command": "Rscript",
      "args": ["/Users/you/apps/mileage/app.R"]
    }
  }
}
```

Clients that can't show apps (Claude Code, which runs in a terminal, is one)
still get every tool the model can call. Their results arrive as text: the
input values and what each output shows.

## Over HTTP

```{r}
serve(app, type = "http", port = 8080)
```

This serves the app at `http://127.0.0.1:8080/mcp`, reachable from this
machine only. To accept connections from elsewhere, listen on all
interfaces with `host = "0.0.0.0"` and put the server behind a proxy that
handles TLS and authentication; shinymcp does neither.

A server listening on this machine refuses requests for host names other
than `localhost` and IP addresses. That protects it from web pages that
point their own domain at your machine (DNS rebinding). If you reach it
through another name, such as a proxy on the same machine, list the name in
`allowed_hosts`. Browser pages from origins other than the server's own
need `allowed_origins`.

The server speaks both generations of the MCP protocol: the versions from
2024-11-05 to 2025-11-25, which start with a handshake and keep a session,
and the stateless 2026-07-28 revision, in which every request stands alone
and any server process can answer it.

## Posit Connect

`mcp_endpoint()` adds an MCP endpoint to a Shiny app. The result is still a
Shiny app, so it deploys like one, to Posit Connect, Shiny Server, or
shinyapps.io:

```{r}
# app.R
library(shiny)
library(shinymcp)

ui <- fluidPage(...)
server <- function(input, output, session) { ... }

shinyApp(ui, server) |>
  mcp_endpoint(name = "mileage", description = "Fuel economy by number of cylinders.")
```

People who open the content's URL get the Shiny app, as before. Chat
clients connect to the same URL followed by `mcp`, for example
`https://connect.example.com/content/1a2b3c4d/mcp`, and get the same app,
live, in the chat. Try it locally with `shiny::runApp()`: the app is at `/`
and the endpoint at `/mcp`.

Given an `McpApp` (or a list of them) instead of a Shiny app,
`mcp_endpoint()` serves the preview page at `/`.

### Signing in

Content that requires a login needs credentials on every request. Clients
send a Connect API key as `Authorization: Key <api-key>`:

```sh
claude mcp add --transport http mileage \
  https://connect.example.com/content/1a2b3c4d/mcp \
  --header "Authorization: Key $CONNECT_API_KEY"
```

Connect checks the key and tells the app who the user is. The server
function sees them as `session$user` and `session$groups`, and tool
functions as `mcp_request()$user` and `mcp_request()$groups`, so an app can
show each person only their data.

### Processes

Each open view of a Shiny app keeps its session in the R process that
started it. If Connect runs the app in several processes, a request can
reach one that doesn't have the session. shinymcp then starts a new session
from the inputs on the page and the app keeps working, but anything the
server function kept outside its inputs starts over. For apps that keep
such state, set **Max processes** to 1 in the content's runtime settings.

Apps built from tools with `mcp_app()` keep no state between calls and can
run in as many processes as you like.

The [`posit-connect`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/posit-connect)
example has a complete app and the client settings for Claude Code and
VS Code.
