---
title: "Hosting MCP Apps in Shiny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hosting MCP Apps in Shiny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Shiny app can show MCP Apps the way a chat client does: in a
conversation built with [shinychat](https://posit-dev.github.io/shinychat/),
where a model opens them, or in a pane anywhere in the app. shinymcp does
the chat client's part. It connects to the app's server, shows the app's
page in a sandboxed frame, and passes the page's requests to the server.

The app can come from anywhere: built with shinymcp in the same R
process, deployed to Posit Connect, a Shiny app served with Shiny's own
MCP support, or an MCP server written in another language.

The examples below host this app:

```{r}
library(shiny)
library(shinymcp)

show_cars <- ellmer::tool(
  function(cyl = "4") {
    cars <- mtcars[mtcars$cyl == as.numeric(cyl), ]
    list(
      scatter = mcp_result_plot(function() plot(cars$wt, cars$mpg)),
      count = paste(nrow(cars), "cars")
    )
  },
  name = "show_cars",
  description = "Plot weight against fuel economy for the cars with a given number of cylinders.",
  arguments = list(
    cyl = ellmer::type_enum(c("4", "6", "8"), "Number of cylinders.", required = FALSE)
  ),
  annotations = ellmer::tool_annotations(read_only_hint = TRUE)
)

cars_app <- mcp_app(
  fluidPage(
    selectInput("cyl", "Cylinders", c(4, 6, 8)),
    plotOutput("scatter"),
    textOutput("count")
  ),
  tools = list(show_cars),
  name = "cars",
  title = "Cars by cylinders"
)
```

## In a chat

`mcp_chat_host()` gives the model in a shinychat conversation the apps'
tools. When it calls one that shows an app, the app appears in the
conversation, live:

```{r}
library(shinychat)

ui <- bslib::page_fillable(chat_ui("chat"))

server <- function(input, output, session) {
  chat <- chat_server("chat", ellmer::chat("anthropic/claude-sonnet-5"))
  mcp_chat_host(chat, cars_app)
}

shinyApp(ui, server)
```

The model gets each tool's structured result, or its text when there is
none, as it would from any MCP client. Then the person uses the app, and
the host keeps the model in step with it:

* **What the person does reaches the model.** An app tells its host what
  the model should know about it; an app built with shinymcp reports the
  values the person has set in it. Before each message the person sends,
  the model is told what every open app reports. It sees the apps as they
  are now, not as they were a few messages ago, and the saved conversation
  keeps only what the person and the model said.
* **Apps can suggest a message.** When an app asks to post one (a button
  such as "Ask about this"), it goes into the chat's input box for the
  person to read and send. `messages = "submit"` sends it at once, for apps
  you trust.
* **Saved conversations come back live.** When shinychat restores a
  conversation, its cards show their apps again, on the result the model
  got, without calling the tools again.

What an app reports comes from the app, so the model is told that, and
each report is cut to 2,000 characters.

The report reaches the model as a message of its own, just before the
person's. A few providers, such as AWS Bedrock, refuse two user messages
in a row. For those, pass `context = FALSE` and add `host$context()` to the
person's message yourself:

```{r}
host <- mcp_chat_host(chat, cars_app, context = FALSE)
host$context() # What the open apps report, or NULL
```

Other arguments set how the cards look: `title` and `icon`, `open` for
whether a card starts expanded, `show_request` for whether it shows the
call's arguments, and `full_screen`. `value_fn` changes what the model
gets.

### Cards alone

`as_shinychat_tool()` is the part of `mcp_chat_host()` that makes the
tools. Register its tools yourself when you want the cards without the
rest:

```{r}
client <- ellmer::chat("anthropic/claude-sonnet-5")
client$register_tool(as_shinychat_tool(cars_app))
chat_server("chat", client)
```

For a remote server, make the tools where the app starts, outside the
server function, and register them in each session. Listing a server's
tools waits for it, so in a session `as_shinychat_tool()` uses only the
list the client already has:

```{r}
sales_tools <- as_shinychat_tool(mcp_client("https://connect.example.com/sales/mcp"))

server <- function(input, output, session) {
  client <- ellmer::chat("anthropic/claude-sonnet-5")
  client$register_tools(sales_tools)
  chat_server("chat", client)
}
```

To put an app in the conversation without a tool call, for example a form
when the chat starts, build the card with `mcp_content_result()` and
append it with `chat_append()`. In a Shiny session `mcp_content_result()`
calls the app's tool and returns a promise, which `chat_append()` waits
for, so the card is saved with what the tool returned and a restored
conversation shows the app as it was.

## In a pane

`mcp_host_ui()` and `mcp_host_server()` host an app anywhere in a Shiny
app, with no model. The host calls the app's tool when the pane opens, with
the `arguments` you give, as a model would:

```{r}
ui <- bslib::page_sidebar(
  sidebar = bslib::sidebar(verbatimTextOutput("context")),
  mcp_host_ui("cars")
)

server <- function(input, output, session) {
  host <- mcp_host_server("cars", cars_app, arguments = list(cyl = "6"))
  output$context <- renderPrint(host$model_context())
}
```

`mcp_host_server()` returns what happens in the app as reactives: the
context it would give a model (`model_context()`), its latest tool call and
result (`last_tool_call()`, `last_result()`), and the messages it asked to
post (`messages()`). `open()` opens the app again with other arguments:

```{r}
observeEvent(input$cylinders, host$open(list(cyl = input$cylinders)))
```

That makes panes useful for reusing an app built for chat clients as part
of a dashboard, for Shiny apps that react to what someone does in one, and
for seeing exactly what a model would be told before you give an app to
one.

For UI created on the server, such as inside `renderUI()`, `mcp_embed()`
does both halves at once.

## Apps from other servers

`mcp_client()` connects to an MCP server over HTTP. Anywhere the examples
above take an app, they also take a client:

```{r}
server <- function(input, output, session) {
  sales <- mcp_client(
    "https://connect.example.com/sales/mcp",
    headers = list(Authorization = paste("Key", Sys.getenv("CONNECT_API_KEY")))
  )
  chat <- chat_server("chat", ellmer::chat("anthropic/claude-sonnet-5"))
  mcp_chat_host(chat, list(cars_app, sales))
}
```

The connection and any credentials stay in R: the app's page never talks to
its server itself, and the browser never sees the key. `headers` can also
be a function, called before each request; create the client in the server
function, as here, for each visitor's requests to carry their own
credentials.

Each source needs a name of its own, since saved conversations refer to
their apps' servers by name. A client is named after its URL; set `name`
to keep it stable if the URL changes.

A Shiny app served with Shiny's own MCP support is a server like any other:
point `mcp_client()` at its `/mcp` path. Shiny apps can hold a request open
while they wait for the app to change, so give the client a `timeout`
longer than that wait.

The client also works on its own, to call a deployed server's tools from
R:

```{r}
sales$tools()
sales$call_tool("open_sales_app", list(region = "West"))
```

## What an app can do

An app's page runs in a sandboxed frame. It can't reach the Shiny page, its
cookies, or the Shiny session, and it can load only what the app declared
(its `csp`); with nothing declared, it has no network access. It can call
its own server's tools that are visible to apps, read its resources, ask to
open links (web and email only), download files, and ask for full screen.
Everything else it asks for is refused.

The model can call only the tools of the sources you give it.

The [`shinychat`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/shinychat),
[`shiny-host`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/shiny-host),
and [`remote-host`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/remote-host)
examples are complete apps.
