---
title: "Apps in shinychat and Shiny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Apps in shinychat and Shiny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

An MCP App doesn't need a desktop chat client. shinymcp can show one inside
a [shinychat](https://posit-dev.github.io/shinychat/) conversation you
build with ellmer, or anywhere in a Shiny app. The app runs the same way in
both: in a sandboxed frame, with its tool calls answered in R. These hosts
show apps made with shinymcp, in the same R process as the Shiny app that
hosts them; they don't connect to other MCP servers.

The examples below host this app:

```{r}
library(shiny)
library(shinymcp)

show_mileage <- ellmer::tool(
  function(cyl = "4") {
    cars <- mtcars[mtcars$cyl == as.numeric(cyl), ]
    list(
      hist = mcp_result_plot(function() hist(cars$mpg, main = NULL, xlab = "Miles per gallon")),
      summary = sprintf("%d cars; median %.1f mpg.", nrow(cars), median(cars$mpg))
    )
  },
  name = "show_mileage",
  description = "Show the fuel economy of the cars in mtcars with a given number of cylinders.",
  arguments = list(
    cyl = ellmer::type_enum(c("4", "6", "8"), "Number of cylinders.", required = FALSE)
  ),
  annotations = ellmer::tool_annotations(read_only_hint = TRUE)
)

app <- mcp_app(
  fluidPage(
    selectInput("cyl", "Cylinders", c(4, 6, 8)),
    plotOutput("hist"),
    textOutput("summary")
  ),
  tools = list(show_mileage),
  name = "mileage"
)
```

## In a shinychat conversation

`as_shinychat_tool()` turns an app's tools into ellmer tools. Register them
with the chat, and when the model calls one, shinychat shows the app, live,
in the tool call's card:

```{r}
library(shinychat)

ui <- bslib::page_fillable(chat_ui("chat"))

server <- function(input, output, session) {
  client <- ellmer::chat("anthropic/claude-sonnet-5")
  client$register_tool(as_shinychat_tool(app))
  chat_server("chat", client)
}

shinyApp(ui, server)
```

The model gets the tool's structured result, or its text when there is
none, as it would from an MCP client, and the person gets the app. What the
person then does in the card stays in the card: ellmer sends the model the
conversation, and shinymcp doesn't add to it. To pass the context an app
publishes on to the model, host the app with `mcp_host_server()` (below),
which reports it.

The card's arguments shape how it looks: `title` and `icon`, `open` for
whether it starts expanded, `show_request` for whether it shows the call's
arguments, and `full_screen`. `value_fn` changes what the model gets, and
`summary` sets the text shown where the app can't be.

To put an app in the conversation without a tool call, for example to open
a form when the chat starts, build the card with `mcp_content_result()` and
append it yourself.

## Anywhere in a Shiny app

`mcp_host_ui()` and `mcp_host_server()` host an app in any Shiny app. The
host calls the app's tool when the app opens, as the model would, with the
`arguments` you give:

```{r}
ui <- bslib::page_sidebar(
  sidebar = bslib::sidebar(verbatimTextOutput("context")),
  mcp_host_ui("mileage")
)

server <- function(input, output, session) {
  host <- mcp_host_server("mileage", app, arguments = list(cyl = "6"))
  output$context <- renderPrint(host$model_context())
}
```

`mcp_host_server()` returns what the app did, as reactives: the context it
would give the model (`model_context()`), its latest tool call and result
(`last_tool_call()`, `last_result()`), and the messages it asked to post to
the chat (`messages()`). `execute()` runs the app's tools with new input
values, and `dispose()` shuts the app down.

That makes the host useful for reviewing an app before you give it to a
model (you see exactly what the model would be told), for dashboards built
around apps, and for Shiny apps that react to what someone does in one.

For UI created on the server, such as inside `renderUI()`, `mcp_embed()`
does both halves at once.

The [`shinychat`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/shinychat)
and [`shiny-host`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/shiny-host)
examples are complete apps.
