---
title: "The MCP Apps Protocol in shinymcp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The MCP Apps Protocol in shinymcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

shinymcp implements the [MCP Apps
specification](https://github.com/modelcontextprotocol/ext-apps)
(version **2026-01-26**), the optional Model Context Protocol extension
that lets MCP servers ship interactive HTML UIs that render inside AI
chat clients. This article explains how the protocol works, exactly
which parts shinymcp implements, and where shinymcp deliberately
deviates. Read this when you want to debug a misbehaving app, integrate
with a new host, or contribute to shinymcp itself.

## The three parties

Every MCP App involves three processes:

1. **Server** — your R process, started by `serve(app)`. It answers
   MCP JSON-RPC requests over stdio or HTTP: `tools/list`,
   `tools/call`, `resources/list`, `resources/read`.
2. **Host** — the chat client (Claude Desktop, ChatGPT, VS Code, or
   shinymcp's own Shiny host and `preview_app()`). It talks MCP to the
   server and renders your app's HTML inside a sandboxed iframe.
3. **View** — your app's HTML page running inside that iframe. The
   shinymcp JS bridge (inlined into the page) talks to the host with
   postMessage/JSON-RPC. It never talks to the server directly; the
   host proxies everything.

## Lifecycle walkthrough

A typical session looks like this:

```
Host  -> Server   initialize (advertises extensions["io.modelcontextprotocol/ui"])
Host  <- Server   capabilities + negotiated protocolVersion
Host  -> Server   tools/list
Host  <- Server   tools with _meta.ui.resourceUri = "ui://your-app"
Model            calls a tool; host sees _meta.ui.resourceUri
Host  -> Server   resources/read "ui://your-app"
Host  <- Server   HTML (text/html;profile=mcp-app) + _meta.ui (CSP, border)
Host             renders HTML in a sandboxed iframe

View  -> Host     ui/initialize (appInfo, appCapabilities)
View  <- Host     hostInfo + hostContext (theme, locale, displayMode, styles)
View  -> Host     ui/notifications/initialized
Host  -> View     ui/notifications/tool-input (the model's arguments)
Host  -> View     ui/notifications/tool-result (when the call completes)

User changes an input in the iframe:
View  -> Host     ui/update-model-context (current input values)
View  -> Host     tools/call (debounced)
Host  -> Server   tools/call (proxied)
View  <- Host     result -> bridge updates output elements

Host  -> View     ui/resource-teardown (before removing the iframe)
View  <- Host     response, then cleanup
```

Two messages in that flow do the heavy lifting for the AI experience:

- **`ui/update-model-context`** pushes the UI's current state into the
  model's context. This is why the model "knows" what the user selected
  in the rendered app and can act on it in the next turn.
- **`ui/notifications/tool-input`** flows the other way: when the model
  calls a tool, the host pushes those arguments into the UI so the
  rendered controls match what the model asked for.

## Compliance summary

### Server (MCP over stdio/HTTP)

| Spec feature | Status |
|---|---|
| `initialize` version negotiation (2024-11-05 … 2025-11-25) | ✅ |
| Extension capability check (`io.modelcontextprotocol/ui`) | ✅ graceful degradation (see below) |
| `tools/list`, `tools/call` | ✅ |
| `resources/list`, `resources/read` | ✅ |
| `ping` | ✅ |
| Tool `_meta.ui.resourceUri` | ✅ (plus deprecated flat `ui/resourceUri` alias) |
| Tool `_meta.ui.visibility` (`"model"` / `"app"` scoping) | ✅ via `mcp_app(tool_visibility = )` |
| Resource `_meta.ui.csp` / `permissions` / `prefersBorder` | ✅ via `mcp_app(csp = , permissions = , prefers_border = )` |
| `tools` `outputSchema` | ✅ generated via `mcp_app(tool_outputs = )`, or pass-through for plain-list tools |
| Extra (non-UI) resources | ✅ via `mcp_app(resources = )` |
| Streamable HTTP session management (`Mcp-Session-Id`) | ✅ assigned on `initialize`, `DELETE` terminates |
| List pagination (`cursor`) | ❌ not needed for single-app servers |
| `resources/subscribe`, `listChanged` notifications | ❌ |

### View (JS bridge, MCP Apps 2026-01-26)

| Spec feature | Status |
|---|---|
| `ui/initialize` handshake + `ui/notifications/initialized` | ✅ |
| `ui/notifications/tool-input` / `tool-result` / `tool-cancelled` / `tool-input-partial` | ✅ (partial input accepted, not yet rendered progressively) |
| `ui/update-model-context` (request semantics) | ✅ |
| `ui/notifications/size-changed` via ResizeObserver | ✅ |
| `ui/resource-teardown` (responds before cleanup) | ✅ |
| `ping` | ✅ |
| JSON-RPC error responses surfaced to callers | ✅ |
| Host context: `theme` → `data-bs-theme`, `locale`, `styles.variables` | ✅ applied automatically |
| `ui/open-link`, `ui/message`, `ui/request-display-mode`, `notifications/message` | ✅ exposed via `window.shinymcp` (below) |
| `resources/read` from the view | ✅ `window.shinymcp.readResource(uri)` |

### Graceful degradation

MCP Apps is an *optional, negotiated* extension. Clients advertise
support in `initialize` under
`capabilities.extensions["io.modelcontextprotocol/ui"]`. When a client
does **not** advertise it, shinymcp still serves all tools — it just
omits the `_meta.ui` annotations. Your tools keep working as ordinary
text tools in any MCP client; the interactive UI lights up only where
the host supports it. You don't need to do anything to get this
behavior.

## Theming: matching the host's light/dark mode

The host tells the view its theme in `hostContext` (and again via
`ui/notifications/host-context-changed` when it changes). The shinymcp
bridge applies it automatically by setting `data-bs-theme="dark"` (or
`"light"`) on the document element:

- **bslib apps** (Bootstrap 5.3+) restyle automatically.
- **Plain shinymcp components** get a built-in dark variant.
- **Custom CSS** can opt in with selectors like
  `:root[data-bs-theme="dark"] .my-widget { ... }`.

Host-provided `styles.variables` (CSS custom properties) are also
applied to `:root`, so hosts that publish design tokens flow through.

## CSP: why external assets fail, and how to declare them

This is the most common production footgun. **Hosts apply a restrictive
Content Security Policy to the iframe** — by default roughly
`default-src 'none'; script-src 'self' 'unsafe-inline'; connect-src 'none'`.
That means:

- No `fetch()`/XHR/WebSocket to any external service.
- No scripts, stylesheets, fonts, or images loaded from CDNs.
- Failures are *silent* — assets just don't load.

shinymcp's default posture avoids all of this: `html_resource()`
**inlines every dependency** (bslib CSS/JS, your styles) into the HTML,
and plots travel as base64 data URIs. If your app stays inside that
model, you never think about CSP.

When you genuinely need external access — say an htmlwidget loading
from a CDN, or client-side calls to an API — declare the domains so
spec-compliant hosts can allow them:

```{r}
app <- mcp_app(
  ui,
  tools,
  name = "my-app",
  csp = list(
    connect_domains = c("https://api.example.com"),
    resource_domains = c("https://cdn.jsdelivr.net")
  )
)
```

These are published as `_meta.ui.csp` on the `ui://` resource. Hosts
MUST NOT allow domains you didn't declare, so be complete. There are
also `frame_domains` (nested iframes) and `base_uri_domains`.

Related declarations:

- `permissions = list(camera = list())` for iframe permissions.
- `prefers_border = TRUE` to hint the host should draw a frame.

## Lazy-loading data with extra resources

Inlining keeps apps CSP-proof, but it also means every byte ships in the
initial HTML resource. For large datasets, declare them as **extra
resources** and fetch them on demand from the iframe — the request flows
through the host to your R process, so no CSP declaration is needed:

```{r}
app <- mcp_app(
  ui,
  tools,
  name = "my-app",
  resources = list(
    "ui://my-app/data" = list(
      content = function() jsonlite::toJSON(big_dataset),
      mime_type = "application/json",
      description = "Full dataset for client-side filtering"
    )
  )
)
```

Function content is evaluated on every read (so it can reflect current
state); plain strings are served as-is. In the app's JS:

```js
window.shinymcp.readResource("ui://my-app/data").then(function (result) {
  var rows = JSON.parse(result.contents[0].text);
  // render client-side
});
```

This works in real chat hosts (which proxy `resources/read` natively), in
shinymcp's Shiny host, and in `preview_app()`.

## Tool output schemas

Tools that return a named list keyed by output ids can publish an
`outputSchema` so hosts and models know the result shape. Declare the
mapping and shinymcp generates the schema, deriving property descriptions
from the matching UI output types (plot outputs are documented as base64
PNGs, tables as HTML markup, and so on):

```{r}
app <- mcp_app(
  ui,
  tools,
  name = "my-app",
  tool_outputs = list(explore = c("scatter", "stats"))
)
```

Only declare tools whose results really match: per the MCP spec, a tool
with an `outputSchema` MUST return conforming `structuredContent`, so
shinymcp never guesses — undeclared tools simply omit the schema.

## Tool visibility: app-only and model-only tools

By default every tool is visible to both the model and the rendered
UI. The spec's `_meta.ui.visibility` lets you scope that, and it maps
naturally onto shinymcp's design where fine-grained recompute tools can
clutter the model's tool list:

```{r}
app <- mcp_app(
  ui,
  tools,
  name = "dashboard",
  tool_visibility = list(
    redraw_plot = "app",     # UI plumbing; hidden from the model
    export_report = "model"  # model-driven; UI must not call it
  )
)
```

## The `window.shinymcp` API

Inside your app's HTML you can script richer host interactions. The
bridge exposes a small global:

```js
// Call a server tool and handle the result yourself
window.shinymcp.callTool("get_data", { region: "EMEA" }).then(...)

// Read a server resource (see "Lazy-loading data" below)
window.shinymcp.readResource("ui://my-app/data").then(function (result) {
  var data = JSON.parse(result.contents[0].text);
});

// Push state into the model's context for future turns
window.shinymcp.updateModelContext({ selected_rows: [1, 5, 8] });

// Ask the host to open a link, send a chat message, or go fullscreen
window.shinymcp.openLink("https://example.com/docs");
window.shinymcp.sendMessage("Please summarize the selected rows.");
window.shinymcp.requestDisplayMode("fullscreen");

// Logging and host info
window.shinymcp.log("info", { event: "filter-applied" });
window.shinymcp.getHostContext(); // { theme, locale, displayMode, ... }
```

All of these are spec methods (`tools/call`, `ui/update-model-context`,
`ui/open-link`, `ui/message`, `ui/request-display-mode`,
`notifications/message`); hosts that don't implement one simply return
an error, which arrives as a rejected Promise.

## Known deviations from the spec

shinymcp aims for strict compliance, with these documented exceptions:

1. **Private host extensions.** shinymcp's bundled Shiny host sends two
   non-spec notifications to drive its Apply/Reset toolbar:
   `ui/notifications/trigger-tool-call` and `ui/notifications/reset`.
   The bridge also accepts `x-shinymcp/trigger-tool-call` and
   `x-shinymcp/reset` spellings; a future release will migrate the
   host to the `x-shinymcp/` namespace to vacate the spec-reserved
   `ui/notifications/*` prefix. Standard hosts never send these, so
   interop is unaffected.
2. **Legacy `_meta["ui/resourceUri"]`.** Tools carry both the current
   nested key and the deprecated flat key for compatibility with hosts
   that haven't updated. The flat key will be dropped once it's removed
   from the spec.
3. **Shiny host sandboxing.** `mcp_embed()` / `mcp_host_server()`
   render the app via `srcdoc` with
   `sandbox="allow-scripts allow-same-origin"` on the *same origin* as
   your Shiny app. This is appropriate for embedding **your own**
   trusted apps; it is not a hardened boundary for third-party HTML.
   The spec's sandbox-proxy pattern (separate origin) is out of scope
   for the built-in host.
4. **`initialArguments` in `hostContext`.** shinymcp's own hosts pass
   initial tool arguments in `hostContext.initialArguments` *in
   addition to* the spec's `ui/notifications/tool-input`, which the
   bridge also handles. Standard hosts only use the latter; both work.

## Debugging

`preview_app(app)` runs a local reference host with a **protocol log
panel** (bottom-right toggle) showing every JSON-RPC message in both
directions — usually the fastest way to see what a real host will see.
For failure modes and fixes, see `vignette("debugging-shinymcp")`.
