---
title: "Rewriting an app as tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rewriting an app as tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`as_mcp_app()` serves a Shiny app as it is, and for most apps that is the
right choice. This article is about the other route: rewriting the app's
reactive code as tools, plain R functions that take the inputs and return
the outputs.

## When to rewrite

A rewrite takes work, so it should buy something. It does when:

* **The model should use the computation on its own.** A tool is useful
  without the app: in chat clients that can't show apps, to other agents,
  or when the model wants the numbers rather than the picture. A Shiny
  app's tool opens the app.
* **Each call should stand alone.** Tools keep no session in R, so any
  process can answer any call, and a restart loses nothing.
* **The app is small.** A few inputs and outputs connected by a reactive
  expression or two rewrite in minutes.

It doesn't when the app is large, keeps state as the person works
(`reactiveVal()`, uploaded data, a multi-step workflow), or relies on
observers with side effects. Serve those as they are.

## A first draft with `convert_app()`

Take this app:

```{r}
ui <- fluidPage(
  selectInput("dataset", "Dataset:", c("mtcars", "iris")),
  numericInput("obs", "Observations:", 10, min = 1, max = 50),
  textOutput("summary_text"),
  tableOutput("data_table")
)

server <- function(input, output, session) {
  selected_data <- reactive({
    head(get(input$dataset, envir = asNamespace("datasets")), input$obs)
  })
  output$summary_text <- renderText({
    paste("Showing", nrow(selected_data()), "rows of", input$dataset)
  })
  output$data_table <- renderTable(selected_data())
}
```

`convert_app()` reads its code (without running it), follows each output
back through the reactive expressions to the inputs, and groups outputs
that share inputs into one tool each. Here both outputs depend on
`selected_data()`, so they become one tool:

```{r}
convert_app("path/to/dashboard")
```

It writes `ui.R`, `tools.R`, and `app.R` to `path/to/dashboard_mcp/`. The
draft tool takes the inputs as typed arguments, with the app's defaults,
and carries the app's code as comments:

```{r}
update_summary_text_and_data_table <- ellmer::tool(
  function(dataset = "mtcars", obs = 10) {
    # From the Shiny app:
    #
    # selected_data <- reactive({
    #     head(get(input$dataset, envir = asNamespace("datasets")), input$obs)
    # })
    #
    # output$summary_text <- renderText({
    #     paste("Showing", nrow(selected_data()), "rows of", input$dataset)
    # })
    #
    # output$data_table <- renderTable({
    #     selected_data()
    # })
    list(
      summary_text = "TODO: output$summary_text",
      data_table = "TODO: output$data_table"
    )
  },
  name = "update_summary_text_and_data_table",
  description = "Update summary_text and data_table based on Dataset, Observations",
  arguments = list(
    dataset = ellmer::type_enum(c("mtcars", "iris"), "Dataset", required = FALSE),
    obs = ellmer::type_number("Observations", required = FALSE)
  ),
  annotations = ellmer::tool_annotations(read_only_hint = TRUE)
)
```

The draft runs as it is, so `preview_app("path/to/dashboard_mcp/app.R")`
shows the app with placeholder outputs straight away.

## Finishing the tools

Rewrite each body as a function of its arguments: reactive expressions
become ordinary variables, `input$x` becomes `x`, and each render call
becomes the value it rendered. Give the tool a name and description that
say what it's for, since those are what the model reads:

```{r}
show_dataset <- ellmer::tool(
  function(dataset = "mtcars", obs = 10) {
    data <- head(getExportedValue("datasets", dataset), obs)
    list(
      summary_text = paste("Showing", nrow(data), "rows of", dataset),
      data_table = data
    )
  },
  name = "show_dataset",
  description = "Show the first rows of one of R's built-in datasets.",
  arguments = list(
    dataset = ellmer::type_enum(c("mtcars", "iris"), "Dataset to show.", required = FALSE),
    obs = ellmer::type_integer("Number of rows, 1 to 50.", required = FALSE)
  ),
  annotations = ellmer::tool_annotations(read_only_hint = TRUE)
)
```

Test it by calling it, as the model would:

```{r}
app$call_tool("show_dataset", list(dataset = "iris", obs = 3))
```

Some things need more than a line-by-line rewrite:

* **Plots.** Return `mcp_result_plot(function() ...)`, or a ggplot object.
* **Observers and side effects.** Code that writes files, sends messages,
  or changes state belongs in its own tool, usually one only the app may
  call (`mcp_app(tool_visibility = list(save = "app"))`). An
  `observeEvent(input$save, ...)` becomes a tool that takes `save`, the
  button's id: it runs when the button is pressed, and not when its other
  inputs change. `CONVERSION_NOTES.md` lists the observers that need this.
* **State kept between interactions.** A tool can't remember, so either
  pass the state in as an argument or keep that part of the app in Shiny
  and serve it with `as_mcp_app()`.

The finished app is in the
[`converted-dashboard`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/converted-dashboard)
example, next to the original.

## Keeping a module's UI

If the app is built from Shiny modules, `mcp_tool_module()` serves a
module's UI with a function of your own in place of its server:

```{r}
mcp_tool_module(
  hist_ui,
  name = "eruptions",
  description = "Histogram of waiting times between Old Faithful eruptions.",
  handler = function(bins = 20) {
    list(plot = mcp_result_plot(function() hist(faithful$waiting, breaks = bins)))
  }
)
```

The handler takes the module's inputs by their ids, without the module's
namespace, and returns its outputs by id.
