---
title: "Rewriting a Shiny app as tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rewriting a Shiny app as tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Shiny app's server function turns inputs into outputs through reactive
expressions. Rewritten as tools, the same code becomes ordinary R functions,
each taking some of the app's inputs as arguments and returning some of its
outputs. The UI stays as it is. `mcp_app()` puts the two together, and the
page calls the tools as the person changes the inputs, much as the server
function would have reacted.

To put an app in a chat without rewriting it, see `vignette("shiny-apps")`.

## When to rewrite

A rewrite takes work, so it should buy something. It does when:

* **The model should use the computation on its own.** A tool is useful
  without the app: in chat clients that can't show apps, to other agents,
  or when the model wants the numbers rather than the picture.
* **Each call should stand alone.** Tools keep no session in R, so any
  process can answer any call, and a restart loses nothing.
* **The model needs only part of the app.** Rewrite that part. The rest
  stays a Shiny app for people to use in the browser.

It doesn't pay when the part the model needs keeps state as the person
works (`reactiveVal()`, uploaded data, a multi-step workflow) or relies on
observers with side effects.

## An app to rewrite

```{r}
ui <- fluidPage(
  titlePanel("Simple Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput("dataset", "Dataset:", c("mtcars", "iris")),
      numericInput("obs", "Observations:", 10, min = 1, max = 50)
    ),
    mainPanel(
      textOutput("summary_text"),
      tableOutput("data_table")
    )
  )
)

server <- function(input, output, session) {
  selected_data <- reactive({
    head(get(input$dataset, envir = asNamespace("datasets")), input$obs)
  })
  output$summary_text <- renderText({
    paste("Showing", nrow(selected_data()), "rows of", input$dataset)
  })
  output$data_table <- renderTable(selected_data())
}
```

## Keep the UI

`mcp_app()` takes the same UI. Tool arguments match inputs by id, and the
names in the list a tool returns match outputs, so `ui` needs no changes.
When an id can't match, because two tools share the page or a module
prefixed it, mark the element with `mcp_input()` or `mcp_output()`.

## Turn the server function into tools

Follow each output back to the inputs it uses. Outputs that use the same
inputs belong in one tool; outputs that use different inputs go in
separate tools, so the page calls only the ones whose inputs changed. Here
both outputs come from `selected_data()`, which uses `dataset` and `obs`,
so one tool returns both.

Then rewrite the code as a function of those inputs:

* `input$x` becomes the argument `x`, with the input's starting value as
  its default;
* a reactive expression becomes a variable, or a helper function when
  several tools use it;
* each render call becomes the value it rendered, in the returned list
  under its output id.

```{r}
show_dataset <- ellmer::tool(
  function(dataset = "mtcars", obs = 10) {
    data <- head(getExportedValue("datasets", dataset), obs)
    list(
      summary_text = paste("Showing", nrow(data), "rows of", dataset),
      data_table = data
    )
  },
  name = "show_dataset",
  description = "Show the first rows of one of R's built-in datasets.",
  arguments = list(
    dataset = ellmer::type_enum(c("mtcars", "iris"), "Dataset to show.", required = FALSE),
    obs = ellmer::type_integer("Number of rows, 1 to 50.", required = FALSE)
  ),
  annotations = ellmer::tool_annotations(read_only_hint = TRUE)
)

app <- mcp_app(ui, tools = list(show_dataset), name = "dashboard")
```

The model reads the tool's name and descriptions, not its code, so say what
it is for. `ellmer::type_enum()` tells the model the choices a select input
offers, and `required = FALSE` lets it leave out arguments that have
defaults.

Each kind of output takes its own kind of value:

| Output | Return |
|---|---|
| `textOutput()`, `verbatimTextOutput()` | a string |
| `tableOutput()` | a data frame |
| `plotOutput()` | a ggplot object, or `mcp_result_plot(function() ...)` for base graphics |
| `uiOutput()`, `htmlOutput()` | an htmltools tag |
| an htmlwidget's output, such as `DT::DTOutput()` | the widget |

Call the tool as the model would, and open the app with `preview_app()`:

```{r}
app$call_tool("show_dataset", list(dataset = "iris", obs = 3))
preview_app(app)
```

The finished app is in the
[`rewritten-dashboard`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/rewritten-dashboard)
example, next to the original.

## What doesn't rewrite line by line

* **Observers and side effects.** Code that writes files, sends messages,
  or changes state belongs in its own tool, usually one only the app may
  call (`mcp_app(tool_visibility = list(save = "app"))`), annotated with
  `ellmer::tool_annotations(read_only_hint = FALSE)`. An
  `observeEvent(input$save, ...)` becomes a tool that takes `save`, the
  button's id: it runs when the button is pressed, and not when its other
  inputs change.
* **State kept between interactions.** A tool can't remember. Pass the
  state in as an argument, or keep that part of the app in Shiny.
* **File uploads.** A tool can't take a file from the page. Take a path or
  the data as an argument, for the model to supply.
* **Inputs made by `renderUI()`.** Inputs in HTML that a tool returns
  aren't connected to tools. Put every input in the UI from the start, and
  show the ones that apply with `conditionalPanel()`.

## Keeping a module's UI

If the app is built from Shiny modules, `mcp_tool_module()` serves a
module's UI with a function of your own in place of its server:

```{r}
mcp_tool_module(
  hist_ui,
  name = "eruptions",
  description = "Histogram of waiting times between Old Faithful eruptions.",
  handler = function(bins = 20) {
    list(plot = mcp_result_plot(function() hist(faithful$waiting, breaks = bins)))
  }
)
```

The handler takes the module's inputs by their ids, without the module's
namespace, and returns its outputs by id.

## With a coding agent

shinymcp comes with a
[skill](https://github.com/JamesHWade/shinymcp/tree/main/inst/skills/convert-shiny-app)
that takes a coding agent through this rewrite. For Claude Code, copy its
folder, `system.file("skills", "convert-shiny-app", package = "shinymcp")`,
into `~/.claude/skills/`.
