---
title: "Get started with shinymcp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with shinymcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("shiny", quietly = TRUE) &&
    requireNamespace("ellmer", quietly = TRUE)
)
```

An MCP App is an interactive page that an AI chat client shows inside the
conversation. The page belongs to an MCP server, which also offers tools the
model can call. When the model calls a tool that has a page, the client
shows the page with the result, and the person can keep working with it.

This article builds a small app from one tool and takes it from your R
session to Claude Desktop.

## A tool and a page

The tool is an R function. The model will call it with arguments, so
describe it and them for the model:

```{r tool, message = FALSE}
library(shiny)
library(shinymcp)

faithful_histogram <- ellmer::tool(
  function(bins = 20) {
    list(
      histogram = mcp_result_plot(
        function() hist(faithful$waiting, breaks = bins, main = NULL, xlab = "Minutes"),
        text = "Histogram of waiting times between eruptions."
      ),
      caption = sprintf(
        "%d eruptions; median wait %d minutes.",
        nrow(faithful),
        median(faithful$waiting)
      )
    )
  },
  name = "faithful_histogram",
  description = "Show a histogram of waiting times between Old Faithful eruptions.",
  arguments = list(
    bins = ellmer::type_integer("Number of bins, 5 to 50.", required = FALSE)
  ),
  annotations = ellmer::tool_annotations(read_only_hint = TRUE)
)
```

The page is Shiny UI. `mcp_app()` puts the two together:

```{r app}
app <- mcp_app(
  ui = fluidPage(
    titlePanel("Old Faithful eruptions"),
    sliderInput("bins", "Number of bins", min = 5, max = 50, value = 20),
    plotOutput("histogram"),
    textOutput("caption")
  ),
  tools = list(faithful_histogram),
  name = "faithful",
  title = "Old Faithful"
)
app
```

Names connect the tool and the page. The argument `bins` takes its value
from the input with id `bins`, and the list the tool returns fills the
outputs `histogram` and `caption`. When the person moves the slider, the
page calls the tool with the new value and draws what it returns. There is
no server function: every call stands alone, and nothing is kept in R
between calls.

## Preview it

```{r preview, eval = FALSE}
preview_app(app)
```

`preview_app()` starts a local MCP server for the app and opens a page that
plays the part of a chat client. It calls the tool the way a model would
and shows the app in the same kind of sandboxed frame a client uses, under
the same content security policy. Next to the app it shows:

* **Model**: what the model received from the call.
* **Context**: what the app told the model after the person changed
  something.
* **Calls**: every tool call, from the model and from the page.
* **Messages**: the protocol messages between the page and the host.
* **Server**: the tools and resources the server offers.

Use the arguments box to call the tool again with other values, and the
theme and width menus to see the app in a dark chat or a narrow one.

## What the model sees

`$run_tool()` returns what an MCP client receives:

```{r run-tool}
result <- app$run_tool("faithful_histogram", list(bins = 30))
cat(result$content[[1]]$text)
```

The text is written for the model, and for clients that can't show the
page. The result also carries:

* `structuredContent`, one value per output:

  ```{r structured}
  str(result$structuredContent)
  ```

* an image of the plot, so a model that reads images can see it;
* in `_meta`, everything the page needs to draw the outputs. Clients pass
  this to the page and keep it from the model.

`$call_tool()` returns the tool's R value instead, which is what you want
in tests.

## Put it in Claude Desktop

A chat client runs your app as a program. Save the app in a file that ends
with `serve()`:

```{r app-file, eval = FALSE}
# ~/apps/faithful/app.R
library(shiny)
library(shinymcp)

faithful_histogram <- ellmer::tool(...)

app <- mcp_app(
  ui = fluidPage(...),
  tools = list(faithful_histogram),
  name = "faithful",
  title = "Old Faithful"
)
serve(app)
```

`serve()` speaks MCP on standard input and output. Test the file in a
terminal first: `Rscript ~/apps/faithful/app.R` should sit waiting for
input (press Ctrl+C to stop it).

Then add it in Claude Desktop under **Settings > Developer > Edit Config**:

```json
{
  "mcpServers": {
    "faithful": {
      "command": "Rscript",
      "args": ["/Users/you/apps/faithful/app.R"]
    }
  }
}
```

Use full paths: Claude Desktop doesn't start in your home directory and, on
macOS, doesn't see your shell's `PATH`, so `Rscript` may need its full path
too (`which Rscript` prints it). Restart Claude Desktop and ask it to "show
the Old Faithful waiting times with 30 bins".

`vignette("deployment")` covers other clients, HTTP, and Posit Connect.

## Next steps

* `vignette("tools")` covers inputs, outputs, and results in more depth,
  and tools only the page may call.
* `vignette("hosting")` puts apps in shinychat conversations and Shiny
  dashboards.
* `vignette("rewriting-as-tools")` turns a Shiny app you already have into
  an app like this one.
