---
title: "shinymcp by example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{shinymcp by example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This is a tour of shinymcp, built one idea at a time.

A Shiny app lives behind a URL. You open it in a browser, change some inputs, read the result, and copy anything useful back to wherever you were working. shinymcp removes that round trip. It turns a Shiny-style app into an MCP App that runs inside an AI assistant like Claude Desktop, so the inputs, plots, and tables appear in the conversation, respond to the user, and run your R code in place.

Read as far as your problem needs and stop. Each section shows the idea with enough code to follow it, links to the full app you can run, and says when to reach for the pattern.

## The one idea behind all of them

An MCP App is two things: some UI, and one or more tools. A tool is an ordinary R function with a name, a description, and typed arguments, written as an `ellmer::tool()`. shinymcp connects the UI to the tools with a single convention based on names. Each tool argument is filled from the input whose `id` matches the argument name, and each name in the list a tool returns is written into the output whose `id` matches it.

That's the whole contract. The wiring is done by a small, self-contained JavaScript bridge, not by Shiny's runtime: the bridge reads the input values, calls your tool, and drops each returned value back into the matching output. Every example below changes one piece of that loop: the inputs, the outputs, where the tool runs, or how you declare it.

## Running the examples

Every app ships with the package. Preview an MCP App in your browser with `preview_app()`, which starts a local host that speaks the MCP Apps protocol, so inputs trigger real tool calls and outputs update live:

```{r}
shinymcp::preview_app(system.file("examples", "hello-mcp-minimal", package = "shinymcp"))
```

A few of the later examples are full Shiny apps rather than bare MCP Apps; run those with `shiny::runApp()`. Each section says which it is and links to the source.

## 1. hello-mcp-minimal: the shape everything builds on

Start with the smallest app that does something, so the structure is easy to see. One input, one tool, one output:

```{r}
library(shinymcp)

ui <- htmltools::tagList(
  mcp_text_input("name", "Your name", value = "world"),
  mcp_text("greeting")
)

tools <- list(
  ellmer::tool(
    fun = function(name = "world") {
      list(greeting = paste0("Hello, ", name, "!"))
    },
    name = "greet",
    description = "Greet a person by name",
    arguments = list(name = ellmer::type_string("The name to greet"))
  )
)

app <- mcp_app(ui, tools, name = "hello-mcp-minimal")
serve(app)
```

The names do the wiring. The tool argument `name` is filled from the input with `id = "name"`. The tool returns `list(greeting = ...)`, and that value goes to the output with `id = "greeting"`. `mcp_app()` bundles the UI and the tools, and `serve()` makes them callable. Run it and type in the box: each keystroke calls `greet()` and updates the greeting.

```{r}
shinymcp::preview_app(system.file("examples", "hello-mcp-minimal", package = "shinymcp"))
```

Reach for this when a user should see and steer the inputs. If you only need a function the model can call with no interface of its own, a plain MCP tool registered directly with ellmer is simpler. The card is the thing worth adding.

Full app: [`inst/examples/hello-mcp-minimal/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/hello-mcp-minimal/app.R).

## 2. hello-mcp: real outputs and a theme

A greeting is a string. Most apps return a plot and a table. Outputs work the same way as inputs: pick the right component, return a value under the matching name. `mcp_plot()` expects a base64-encoded PNG, and `mcp_text()` expects a string. Wrapping the UI in a bslib `page()` lets the card inherit a design system.

```{r}
ui <- bslib::page(
  theme = bslib::bs_theme(preset = "shiny"),
  bslib::card(
    bslib::card_header("Dataset Explorer"),
    bslib::layout_columns(
      col_widths = c(4, 8),
      mcp_select("dataset", "Choose dataset", c("mtcars", "iris", "pressure")),
      htmltools::tagList(
        mcp_plot("plot", height = "280px"),
        mcp_text("summary")
      )
    )
  )
)

get_summary <- ellmer::tool(
  fun = function(dataset = "mtcars") {
    data <- get(dataset, envir = asNamespace("datasets"))

    tmp <- tempfile(fileext = ".png")
    grDevices::png(tmp, width = 600, height = 280, res = 96)
    on.exit(unlink(tmp))
    plot(data, main = dataset)  # any base plot works
    grDevices::dev.off()

    list(
      summary = paste(capture.output(summary(data)), collapse = "\n"),
      plot = base64enc::base64encode(tmp)
    )
  },
  name = "get_summary",
  description = "Summary statistics and a plot for the selected dataset",
  arguments = list(dataset = ellmer::type_string("Dataset name"))
)
```

Returning a plot always works the same way: open a graphics device pointed at a temp file, draw, then return `base64enc::base64encode()` of the file under the plot output's name. The bridge knows an `mcp_plot()` output takes an image, so it renders the string as one.

```{r}
shinymcp::preview_app(system.file("examples", "hello-mcp", package = "shinymcp"))
```

If the answer is a single number or a sentence, skip the plot. `mcp_text()` on its own is the right amount of machinery.

Full app: [`inst/examples/hello-mcp/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/hello-mcp/app.R).

## 3. penguins: native shiny and bslib inputs

So far the inputs have used shinymcp's own `mcp_select()` and `mcp_text_input()`. You don't have to. The bridge matches tool argument names to element `id`s in the rendered page, and a standard `shiny::selectInput("species", ...)` produces an element with `id = "species"`. Your existing inputs work untouched, as long as the ids line up with the argument names.

```{r}
ui <- bslib::page_sidebar(
  theme = bslib::bs_theme(preset = "shiny"),
  title = "Palmer Penguins Explorer",
  sidebar = bslib::sidebar(
    shiny::selectInput("species", "Species", c("All", "Adelie", "Chinstrap", "Gentoo")),
    shiny::selectInput("x_var", "X axis", var_choices),
    shiny::selectInput("y_var", "Y axis", var_choices, selected = "bill_depth_mm"),
    shiny::checkboxInput("trend", "Show trend line")
  ),
  bslib::card(bslib::card_header("Scatter Plot"), mcp_plot("scatter", height = "380px")),
  bslib::card(bslib::card_header("Summary Statistics"), mcp_text("stats"))
)

explore_penguins <- ellmer::tool(
  fun = function(species = "All", x_var = "bill_length_mm",
                 y_var = "bill_depth_mm", trend = FALSE) {
    # ... build a ggplot, save to PNG, summarise ...
    list(scatter = plot_b64, stats = stats)
  },
  name = "explore_penguins",
  description = "Explore the Palmer Penguins dataset",
  arguments = list(
    species = ellmer::type_string("Species filter"),
    x_var = ellmer::type_string("X axis variable name"),
    y_var = ellmer::type_string("Y axis variable name"),
    trend = ellmer::type_boolean("Whether to show a linear trend line")
  )
)
```

The four arguments `species`, `x_var`, `y_var`, and `trend` are filled from the four inputs with the same ids. `trend` is a `type_boolean()`, so the checkbox arrives as `TRUE` or `FALSE`. This is the easiest route for an app you already have: keep the inputs, write the tool, line up the names.

```{r}
shinymcp::preview_app(system.file("examples", "penguins", package = "shinymcp"))
```

The `mcp_*()` inputs are still worth using when you want the contract spelled out on the page, or when a widget's id won't match the name you want. That case is next.

Full app: [`inst/examples/penguins/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/penguins/app.R).

## 4. bslib-inputs: escape hatches for awkward ids

Auto-detection depends on the id matching the argument name. Sometimes it won't. `radioButtons()` puts its `id` on a wrapping `<div>`, not on the inputs the bridge reads. For those cases, two functions let you point the bridge by hand: `mcp_input()` stamps the attribute the bridge looks for onto any tag, and `mcp_output()` marks any tag as an output target.

```{r}
sidebar <- bslib::sidebar(
  shiny::selectInput("dataset", "Dataset", c("mtcars", "iris", "pressure")),
  shiny::numericInput("n", "Rows to show", value = 10, min = 1, max = 50),

  # radioButtons puts the id on a wrapper, so name the input explicitly
  mcp_input(
    shiny::radioButtons("format", "Format", c("summary", "head")),
    id = "format"
  )
)

# turn any tag into an output target
result <- mcp_output(tags$pre(id = "result", style = "white-space: pre-wrap;"))
```

Both are manual overrides. The first two inputs need nothing; the radio group needs `mcp_input()` because its structure hides the id. Use the override only where auto-detection fails. Sprinkling it everywhere just adds noise.

```{r}
shinymcp::preview_app(system.file("examples", "bslib-inputs", package = "shinymcp"))
```

Full app: [`inst/examples/bslib-inputs/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/bslib-inputs/app.R).

## 5. bind-mcp-demo: expose part of an app you already have

The examples so far build an MCP App from scratch. Often you have a working Shiny app and want to expose only part of it to an agent, without rewriting it or breaking the version your colleagues already use. `bindMcp()` is a pipe-friendly annotation: add it to an input or output tag to make that element visible to MCP. Anything you don't annotate keeps working in Shiny but stays invisible to the agent.

```{r}
ui <- bslib::page_sidebar(
  sidebar = bslib::sidebar(
    # exposed to the agent
    selectInput("dataset", "Dataset:", c("mtcars", "iris", "pressure")) |> bindMcp(),
    # works in Shiny, hidden from the agent
    numericInput("n", "Rows to show:", value = 10, min = 1, max = 50)
  ),
  bslib::card(
    plotOutput("plot") |> bindMcp(),
    verbatimTextOutput("summary") |> bindMcp()
  )
)

# the ordinary server() is unchanged; wrap the running app and attach a tool
app <- shinyApp(ui, server) |>
  as_mcp_app(
    name = "dataset-explorer",
    tools = list(
      ellmer::tool(
        fun = function(dataset = "mtcars") list(summary = ..., plot = ...),
        name = "explore_dataset",
        description = "Explore a dataset",
        arguments = list(dataset = ellmer::type_string("Dataset name"))
      )
    )
  )
serve(app)
```

`bindMcp()` decides the boundary, and `as_mcp_app()` wraps the existing `shinyApp()` and gives it the tool handler the agent calls. The `n` input is left unbound on purpose: exposure is a choice you make element by element, which matters when some controls shouldn't be in the agent's reach.

```{r}
shinymcp::preview_app(system.file("examples", "bind-mcp-demo", package = "shinymcp"))
```

Use this for incremental adoption. When you're building something new, `mcp_app()` keeps the contract explicit from the first line, which is usually clearer.

Full app: [`inst/examples/bind-mcp-demo/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/bind-mcp-demo/app.R).

## 6. multi-tool: more than one tool in an app

One app can do two unrelated jobs, and they should be two tools. If a data explorer and a greeting box shared a single tool, changing your name would re-run the data query. shinymcp avoids that by working out which inputs and outputs are connected and giving each connected group its own tool. The bridge then routes an input change only to the tool that owns it.

```{r}
explore_data <- ellmer::tool(
  fun = function(dataset, n_rows, sort_col) {
    # ... filter, sort, plot ...
    list(summary = ..., plot = ...)
  },
  name = "update_summary_and_plot",
  description = "Explore a dataset: filter rows, sort, and visualise",
  arguments = list(
    dataset = ellmer::type_string("Dataset name"),
    n_rows = ellmer::type_number("Number of rows to show"),
    sort_col = ellmer::type_string("Column to sort by")
  ),
  annotations = ellmer::tool_annotations(read_only_hint = TRUE, idempotent_hint = TRUE)
)

greet <- ellmer::tool(
  fun = function(user_name) list(greeting = paste("Hello,", user_name)),
  name = "update_greeting",
  description = "Generate a personalised greeting",
  arguments = list(user_name = ellmer::type_string("User's name"))
)

app <- mcp_app(ui, tools = list(explore_data, greet), name = "multi-tool-demo")
```

You pass a list of tools, and nothing else changes. The analyzer also follows chained reactives, so if a filtered table depends on a base table that depends on `input$dataset`, the tool still lists `dataset` as an argument even though no single line names it directly. The `tool_annotations()` are hints to the client: here, that the tool only reads data and is safe to call repeatedly.

```{r}
shinymcp::preview_app(system.file("examples", "multi-tool", package = "shinymcp"))
```

Keep one task per tool. If you're tempted to fold two unrelated operations into a single call, make two tools instead.

Full app: [`inst/examples/multi-tool/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/multi-tool/app.R).

## 7. module-tool: reuse a Shiny module

If you've already written a Shiny module, a paired `ui` and `server` function you use in several apps, you can serve it as an MCP App without rewriting it. `mcp_tool_module()` takes the module and a `handler`: the module renders the usual Shiny way, and the handler is the MCP-side function that computes the structured result the client receives.

```{r}
app <- mcp_tool_module(
  module_ui = hist_ui,
  module_server = hist_server,
  name = "histogram",
  description = "Show an interactive histogram with adjustable bins",
  handler = function(dataset = "faithful", bins = 25) {
    data <- if (dataset == "faithful") faithful$eruptions else mtcars$mpg
    tmp <- tempfile(fileext = ".png")
    grDevices::png(tmp, width = 600, height = 280, res = 96)
    on.exit(unlink(tmp))
    hist(data, breaks = as.integer(bins), main = paste(dataset, "histogram"))
    grDevices::dev.off()
    list(plot = base64enc::base64encode(tmp))
  },
  arguments = list(
    dataset = ellmer::type_string("Dataset: 'faithful' or 'mtcars'"),
    bins = ellmer::type_number("Number of histogram bins")
  )
)
```

The same module also drops into a shinychat interface through `shinychat::chat_tool_module()`, so you write the histogram once and reuse it in a dashboard, a chat, and an MCP App. A one-off app doesn't need this; modules earn their keep through reuse.

```{r}
shinymcp::preview_app(system.file("examples", "module-tool", package = "shinymcp"))
```

Full app: [`inst/examples/module-tool/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/module-tool/app.R).

## 8. converted-dashboard: let the converter write the first draft

You don't always have to write the MCP App by hand. Point `convert_app()` at a plain Shiny app and it produces a working starting point. It parses the source with `parse_shiny_app()`, traces the reactive graph with `analyze_reactive_graph()`, and writes the new app with `generate_mcp_app()`. This example keeps both files: `original-app.R` is the input, `app.R` is the generated result.

The input is a plain `fluidPage` dashboard:

```{r}
# original-app.R
ui <- fluidPage(
  titlePanel("Simple Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput("dataset", "Dataset:", c("mtcars", "iris")),
      numericInput("obs", "Observations:", 10, min = 1, max = 50)
    ),
    mainPanel(textOutput("summary_text"), tableOutput("data_table"))
  )
)
```

The generated output uses shinymcp components, with the inputs and outputs mapped across:

```{r}
# app.R (generated)
ui <- htmltools::tagList(
  mcp_select("dataset", "Dataset:", c("mtcars", "iris")),
  mcp_numeric_input("obs", "Observations:", value = 10, min = 1, max = 50),
  mcp_text("summary_text"),
  mcp_table("data_table")
)
```

The converter reads structure, not meaning. It can see that `summary_text` depends on `dataset` and `obs`, but it can't know what your numbers are supposed to be. Treat the result as a first draft and review the generated tool body before you ship it. The [automatic conversion](https://jameshwade.github.io/shinymcp/articles/automatic-conversion.html) article covers the full workflow.

```{r}
shinymcp::preview_app(system.file("examples", "converted-dashboard", package = "shinymcp"))
```

Full app: [`inst/examples/converted-dashboard/`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/converted-dashboard).

## 9. serve-to-client: call your tool from a real assistant

Previewing runs the app on your own machine. To call your R tool from an assistant you already use, serve it. `serve(app, type = "stdio")` turns the app into a server that a client launches as a subprocess and talks to over standard input and output.

```{r}
app <- mcp_app(ui = ..., tools = list(...), name = "shinymcp-demo")
serve(app, type = "stdio")
```

You register that script in the client's configuration. For Claude Desktop, add an entry under `mcpServers` pointing at the file:

```json
{
  "mcpServers": {
    "shinymcp-demo": {
      "command": "Rscript",
      "args": ["/ABSOLUTE/PATH/TO/serve-to-client/serve.R"]
    }
  }
}
```

Restart the client and the tool appears in its tool list; ask it to greet someone and the card renders inline. The example's [README](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/serve-to-client/README.md) gives the exact steps for both Claude Desktop and VS Code, including where each one keeps its config.

While you're still designing the tool, `preview_app()` is the faster loop. Switch to stdio once the contract is settled and you want to test the end-to-end experience.

Full example: [`inst/examples/serve-to-client/`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/serve-to-client).

## 10. shinychat-card: render tool calls as cards in a chat

If you're building a chat interface in Shiny with [shinychat](https://posit-dev.github.io/shinychat/), `as_shinychat_tool()` turns an MCP App into a tool the model can call and renders the result as a card in the conversation.

```{r}
card_app <- mcp_app(
  ui = htmltools::tagList(
    mcp_select("dataset", "Dataset", c("mtcars", "iris", "pressure")),
    mcp_text("summary"),
    mcp_plot("plot", height = "220px")
  ),
  tools = list(list(
    name = "inspect_dataset",
    description = "Inspect a small built-in dataset",
    fun = function(dataset = "mtcars") {
      data <- get(dataset, envir = asNamespace("datasets"))
      list(
        summary = paste(capture.output(summary(data)), collapse = "\n"),
        plot = mcp_result_plot(function() plot(data), text = paste("Scatter for", dataset))
      )
    }
  )),
  name = "shinychat-card-demo"
)

inspect_dataset <- as_shinychat_tool(
  card_app,
  value_fn = function(raw_result) list(summary = raw_result$summary),
  summary = function(raw_result) raw_result$summary,
  title = "Dataset Inspector"
)

# in the server:
client <- ellmer::chat("openai/gpt-4.1-nano")
client$register_tool(inspect_dataset)
```

`mcp_result_plot()` packages the image together with alt text, so the result stays meaningful to a client that can't show pictures. `value_fn` and `summary` separate what the model reads back, a compact summary it can reason about, from what the person sees, the full card. That separation matters once a model is acting on the result.

```{r}
shiny::runApp(system.file("examples", "shinychat-card", package = "shinymcp"))
```

This one is only for shinychat apps. Outside that setting, serving the tool or hosting it in Shiny (next) is enough.

Full app: [`inst/examples/shinychat-card/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/shinychat-card/app.R).

## 11. embed-in-shiny: Shiny as the review surface

Sometimes you want a person to see exactly what the agent would see, inside a familiar dashboard. `mcp_host_ui()` and `mcp_host_server()` embed any `mcp_app()` object into a Shiny app as a sandboxed iframe that runs the real tool. You define the app once and use the same object two ways: served to a client, or hosted here for review.

```{r}
greet_app <- mcp_app(ui = ..., tools = list(...), name = "greet")

ui <- bslib::page_sidebar(
  title = "An MCP App embedded in Shiny",
  sidebar = bslib::sidebar("The card on the right is the same MCP App you would serve to a client."),
  bslib::card(bslib::card_header("Embedded MCP App"), mcp_host_ui("greet"))
)

server <- function(input, output, session) {
  mcp_host_server("greet", greet_app, trigger = "change")
}

shinyApp(ui, server)
```

The `trigger` argument decides when the card calls its tool. `"change"` calls on every edit, which is good for a fast preview; `"submit"` waits for an explicit apply, which is good when a call is expensive or a reviewer wants to control it. Because the host runs the tool, the parent app can read the structured result and react to it.

```{r}
shiny::runApp(system.file("examples", "embed-in-shiny", package = "shinymcp"))
```

For a deployment that talks only to an MCP client, you don't need a host; `serve()` from example 9 is enough. The host is for review surfaces and dashboards where a human stays in the loop.

Full app: [`inst/examples/embed-in-shiny/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/embed-in-shiny/app.R).

## 12. data-explorer: build the inputs from the data

When a data frame's columns drive the controls, computing the choices beats hard-coding them. The id-matches-argument rule from example 3 still holds; only the source of the choices changes.

```{r}
numeric_cols <- names(data)[vapply(data, is.numeric, logical(1))]
categorical_cols <- names(data)[vapply(
  data, function(x) is.character(x) || is.factor(x) || length(unique(x)) <= 10,
  logical(1)
)]

num_choices <- setNames(numeric_cols, gsub("_", " ", numeric_cols))

sidebar <- bslib::sidebar(
  shiny::selectInput("x_var", "X axis", num_choices),
  shiny::selectInput("y_var", "Y axis", num_choices, selected = numeric_cols[2]),
  shiny::selectInput("color_var", "Color by", c("None" = "none", categorical_cols)),
  shiny::selectInput("geom", "Plot type", c("point", "line", "bar", "boxplot", "histogram"))
)
```

The tool takes `x_var`, `y_var`, `color_var`, and `geom` as plain strings and builds the plot to match. Swap in a different frame at the top and the whole UI re-derives itself. That flexibility earns its extra code when the columns genuinely vary between deployments or users; for a fixed, well-known schema, hard-coded inputs read more clearly.

```{r}
shinymcp::preview_app(system.file("examples", "data-explorer", package = "shinymcp"))
```

Full app: [`inst/examples/data-explorer/app.R`](https://github.com/JamesHWade/shinymcp/blob/main/inst/examples/data-explorer/app.R).

## 13. rpharma-hangout: the capstone

The R/Pharma demo puts the pieces together into a realistic, governed app. Two skill cards, a Safety Signal Scout and an Enrollment Rescue Simulator, run inside a clinical dashboard over synthetic trial data. Each skill is an ordinary `mcp_app()`, reachable three ways: served to an MCP client, hosted in the dashboard with `mcp_host_server()`, and registered with shinychat through `as_shinychat_tool()`.

This example adds the handoff. Alongside the display content, each result carries a structured `model_value` that the parent app reads directly, instead of scraping the rendered card:

```{r}
list(
  memo = mcp_result_text(
    memo,
    model_value = list(
      widget = "Safety Signal Scout",
      decision = decision,
      cohort = cohort,
      risk_ratio = rr
    )
  ),
  evidence = mcp_result_table(evidence, text = "Aggregate AE evidence table."),
  risk_plot = mcp_result_plot(draw_plot, model_value = rate_by_arm, text = "Event rate by arm.")
)
```

The result stays aggregate: a decision, a risk ratio, an evidence table, never subject-level rows. The parent reads the typed `model_value` from `mcp_host_server()$last_raw_result()`, so it reacts to data instead of parsing the screen. A contract inspector panel shows the tool's argument schema live, so a reviewer can see exactly what the agent is allowed to call.

```{r}
shiny::runApp(system.file("examples", "rpharma-hangout", package = "shinymcp"))
```

The data is synthetic. The pattern it shows, approved skills with inspectable contracts and aggregate handoffs, applies to any regulated or high-stakes setting where someone needs to see exactly what the agent called and what it got back. When you don't need that oversight and the app covers a single skill, a lone `mcp_app()` served directly (example 9) is far less to build.

Full example and walkthrough: [`inst/examples/rpharma-hangout/`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/rpharma-hangout).

## Going further

These examples keep an LLM out of the loop so they run offline. Two related packages build on the same R functions:

- [`{dsprrr}`](https://jameshwade.github.io/dsprrr/) fits prompts to data with DSPy-style signatures, optimization, and tracing, so you improve a prompt with examples instead of by hand.
- [`{deputy}`](https://jameshwade.github.io/deputy/) is an agent runtime: it lets a tool-using agent call reviewed R functions, with permissions, hooks, and multi-agent coordination.

Together with shinymcp, they treat Shiny components as building blocks for agent-connected workflows rather than standalone dashboards.
