---
title: "Get started with shinymcp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with shinymcp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("shiny", quietly = TRUE)
)
```

An MCP App is an interactive page that an AI chat client shows inside the
conversation. The page belongs to an MCP server, which also offers tools the
model can call. When the model calls a tool that has a page, the client
shows the page with the result, and the person can keep working with it.

This article takes a small Shiny app from your R session to Claude Desktop.

## An app to start with

Any Shiny app will do. Here is a small one:

```{r app, message = FALSE}
library(shiny)
library(shinymcp)

ui <- fluidPage(
  titlePanel("Old Faithful eruptions"),
  sliderInput("bins", "Number of bins", min = 5, max = 50, value = 20),
  plotOutput("histogram"),
  textOutput("caption")
)

server <- function(input, output, session) {
  output$histogram <- renderPlot({
    hist(faithful$waiting, breaks = input$bins, main = NULL, xlab = "Minutes")
  })
  output$caption <- renderText({
    sprintf("%d eruptions; median wait %d minutes.", nrow(faithful), median(faithful$waiting))
  })
}
```

`as_mcp_app()` turns it into an MCP App without changes to the app:

```{r as-mcp-app}
app <- as_mcp_app(
  shinyApp(ui, server),
  name = "faithful",
  title = "Old Faithful",
  description = "Show a histogram of waiting times between Old Faithful eruptions."
)
app
```

The app has two tools. `faithful` is for the model; its arguments are the
app's inputs. `faithful_view` is for the page only: it carries the person's
changes to the app's session in R.

## Preview it

```{r preview, eval = FALSE}
preview_app(app)
```

`preview_app()` starts a local MCP server for the app and opens a page that
plays the part of a chat client. It calls `faithful` the way a model would
and shows the app in the same kind of sandboxed frame a client uses, under
the same content security policy. Next to the app it shows:

* **Model**: what the model received from the call.
* **Context**: what the app told the model after the person changed
  something.
* **Calls**: every tool call, from the model and from the page.
* **Messages**: the protocol messages between the page and the host.
* **Server**: the tools and resources the server offers.

Use the arguments box to call the tool again with other values, and the
theme and width menus to see the app in a dark chat or a narrow one.

## What the model sees

You can call the model's tool from R to see the same result:

```{r run-tool}
result <- app$run_tool("faithful", list(bins = 30))
cat(result$content[[1]]$text)
```

The text is written for the model. The result also carries:

* `structuredContent`, the same facts as data:

  ```{r structured}
  str(result$structuredContent)
  ```

* an image of the plot, so a model that reads images can see it;
* in `_meta`, everything the page needs to draw the outputs. Clients pass
  this to the page and keep it from the model.

The `view` field identifies this view of the app. Every time the model calls
`faithful`, a new view opens with its own session. To change a view that is
already open, the model passes its `view` id back with the new values.

## Put it in Claude Desktop

A chat client runs your app as a program. Save the app in a file that ends
with `serve()`:

```{r app-file, eval = FALSE}
# ~/apps/faithful/app.R
library(shiny)
library(shinymcp)

ui <- fluidPage(...)
server <- function(input, output, session) { ... }

app <- as_mcp_app(shinyApp(ui, server), name = "faithful", title = "Old Faithful")
serve(app)
```

`serve()` speaks MCP on standard input and output. Test the file in a
terminal first: `Rscript ~/apps/faithful/app.R` should sit waiting for
input (press Ctrl+C to stop it).

Then add it in Claude Desktop under **Settings > Developer > Edit Config**:

```json
{
  "mcpServers": {
    "faithful": {
      "command": "Rscript",
      "args": ["/Users/you/apps/faithful/app.R"]
    }
  }
}
```

Use full paths: Claude Desktop doesn't start in your home directory and, on
macOS, doesn't see your shell's `PATH`, so `Rscript` may need its full path
too (`which Rscript` prints it). Restart Claude Desktop and ask it to "show
the Old Faithful waiting times with 30 bins".

`vignette("deployment")` covers other clients, HTTP, and Posit Connect.

## Next steps

* `vignette("shiny-apps")` explains what runs where when you serve a Shiny
  app, and how to choose what the model sees and knows.
* `vignette("tools")` shows the other way to build an app: from tools, with
  no Shiny server function.
* `vignette("hosting")` puts apps in shinychat conversations and Shiny
  dashboards.
