---
title: "Serving a Shiny app as it is"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Serving a Shiny app as it is}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("shiny", quietly = TRUE)
)
```

`as_mcp_app()` serves a Shiny app you already have as an MCP App, without
changes: its server function keeps running in R. This article covers what
happens when the model opens the app and while someone uses it, how to
decide what the model sees and knows, and what the page can do.

Shiny is gaining MCP support of its own
([rstudio/shiny#4407](https://github.com/rstudio/shiny/pull/4407)), which
runs Shiny's own JavaScript against a real session, where shinymcp has
stand-ins for both. Once it is released, it will be the way to serve a live
app, and shinymcp will stop doing so; this article will then cover moving
an app over. shinymcp will still show such apps in Shiny apps and shinychat
conversations, through `mcp_client()`; see `vignette("hosting")`. Until
then, `as_mcp_app()` works as described here, and it
gets bug fixes but no new features. For something the model should be
able to use on its own, rewrite that part of the app as tools instead; see
`vignette("rewriting-as-tools")`.

```{r setup, message = FALSE}
library(shiny)
library(shinymcp)

ui <- fluidPage(
  selectInput("cyl", "Cylinders", c(4, 6, 8)),
  sliderInput("bins", "Bins", min = 5, max = 30, value = 10),
  plotOutput("hist"),
  textOutput("summary")
)

server <- function(input, output, session) {
  cars <- reactive(mtcars[mtcars$cyl == input$cyl, ])
  output$hist <- renderPlot({
    hist(cars()$mpg, breaks = input$bins, main = NULL, xlab = "Miles per gallon")
  })
  output$summary <- renderText({
    sprintf("%d cars with %s cylinders; median %.1f mpg.", nrow(cars()), input$cyl, median(cars()$mpg))
  })
}

app <- as_mcp_app(
  shinyApp(ui, server),
  name = "mileage",
  description = "Fuel economy of the cars in mtcars, by number of cylinders."
)
app
```

## Views and sessions

The app has two tools. The model calls `mileage`. Each call opens a *view*:
shinymcp starts a session of the app's server function, sets its inputs
from the call's arguments, lets it run, and returns what the outputs show.
The chat client then shows the page, and from there the person using it
works with the app directly. Their changes go to that view's session
through `mileage_view`, a tool only the page can call, and the outputs that
changed come back.

So the server function runs as it would for a browser: reactive expressions
cache, observers fire, `updateSelectInput()` and its relatives change the
page, and `validate()` and `req()` behave as usual. There is one session per
view, so two views of the same app don't share state, as two browser tabs
wouldn't.

When the model calls `mileage` again, it opens another view. To change a
view that is already open, it passes that view's id as `view`; the person
then sees the same view update, and inputs the model leaves out keep their
current values.

## What the model sees

The model's tool takes the app's inputs as arguments, described from the
UI: a select input becomes a list of allowed values, a slider a bounded
number, a date input a date.

```{r schema}
str(app$tool_definitions()[[1]]$inputSchema$properties$bins)
```

When it calls the tool, the result starts with text written for the model,
and carries the same facts as structured data:

```{r result}
result <- app$run_tool("mileage", list(cyl = "6", bins = 5))
cat(result$content[[1]]$text)
str(result$structuredContent$outputs)
```

Plots also go to the model as images, if it can read them. Set
`images = FALSE` in `as_mcp_app()` to leave them out and save tokens.

Everything the page needs to draw the outputs travels in the result's
`_meta`, which chat clients pass to the page and keep from the model.

Write a `description` that says what the app is for. The model reads it
when it decides whether to open the app, and a sentence about the purpose
("fuel economy by number of cylinders") helps more than the list of inputs
shinymcp writes by default.

### Choosing inputs and outputs

A large app makes for a tool with many arguments, most of them beside the
point. Mark the ones that matter with `bindMcp()`:

```{r bind}
ui <- fluidPage(
  selectInput("cyl", "Cylinders", c(4, 6, 8)) |> bindMcp(),
  sliderInput("bins", "Bins", min = 5, max = 30, value = 10),
  plotOutput("hist"),
  textOutput("summary") |> bindMcp()
)
marked <- as_mcp_app(shinyApp(ui, server), name = "mileage")
names(marked$tool_definitions()[[1]]$inputSchema$properties)
```

Once anything is marked, only marked inputs are arguments and only marked
outputs are reported. The person using the app still sees and uses all of
it. Action buttons are only for the model to press when they are marked,
and password and file inputs never are.

## What the model knows afterwards

When the person changes something, the page tells the model on its next
turn: by default, the current values of the inputs it can set and a short
summary of each output it reads. To say something more useful, call
`mcp_model_context()` from the server function:

```{r model-context, eval = FALSE}
server <- function(input, output, session) {
  cars <- reactive(mtcars[mtcars$cyl == input$cyl, ])
  observe({
    mcp_model_context(
      text = sprintf("The user is looking at the %d cars with %s cylinders.", nrow(cars()), input$cyl),
      data = list(cyl = input$cyl, cars = rownames(cars()))
    )
  })
}
```

Each call replaces the last. Keep it short and factual: the model reads it
along with everything else in the conversation. `mcp_send_message()` goes
further and posts a message into the chat as the user.

Both do nothing in an ordinary Shiny session, so the same app still runs
with `shiny::runApp()`; `is_mcp_session()` tells the two apart.

## The chat around the app

`mcp_host_context()` returns what the chat client said about itself: its
color theme, whether the app is shown inline or full screen, and the user's
locale and time zone. It is reactive, so a plot that reads the theme redraws
when the user switches the chat to dark mode.

`mcp_request()` describes the call being handled. On Posit Connect it
includes the signed-in user and their groups, which the server function can
use to decide what data to show.

## What works on the page

The page is the app's UI, rendered once to HTML, with shinymcp's JavaScript
in place of Shiny's. It draws Shiny's inputs itself, and shows every kind of
output: text, tables, plots and images, `renderUI()`, and htmlwidgets such
as plotly, DT, reactable, and leaflet. Notifications, modal dialogs,
`insertUI()`, tab changes (`updateTabsetPanel()`, `insertTab()`,
`hideTab()`), downloads, and file uploads work.

`invalidateLater()` and `reactivePoll()` keep running while the app is
open, and an `ExtendedTask`'s result appears when the task finishes.
Clicks, double clicks, hovers, and brushes on a plot
(`plotOutput(click =, brush =)`) send what they send in a browser, so
`nearPoints()` and `brushedPoints()` work.

Select inputs are the browser's own. One whose choices stay on the server
(`updateSelectizeInput(server = TRUE)`) shows the first 1,000 of them,
with a search box for the rest.

`conditionalPanel()` works too, with one difference. Shiny runs a panel's
condition as JavaScript, which chat clients forbid, so shinymcp reads it
instead. It understands what conditions are usually made of: comparisons,
`&&`, `||`, `!`, arithmetic, `input.x` and `output.x` (including an output
set to a `reactive()`), `.length`, regular expressions, and common string
and array methods such as `indexOf()` and `includes()`. A condition it
can't read shows its panel, with a warning in the browser's console.

Packages that talk to Shiny's JavaScript work too. The page provides
`window.Shiny` with what they use: input bindings (shinyWidgets' pickers,
switches, and button groups; bslib's sidebars, accordions, and task
buttons), `Shiny.setInputValue()` (row selection in DT,
`plotly::event_data()`, clicks on a leaflet map), custom message handlers
(shinyjs), and the events that loading spinners listen for
(shinycssloaders).

The app starts the way `shiny::runApp()` starts it: `global.R`, the files
in `R/`, and `onStart` run once, before the UI is built, and the app's code
runs in its own directory. Scripts, stylesheets, and images the UI loads
from `www/`, or from a path added with `addResourcePath()`, are written
into the page.

Some things don't carry over:

* **Content from other sites.** Chat clients show the page under a strict
  Content Security Policy that blocks anything from another domain unless
  the app declares it. Map tiles, web fonts, and scripts from a CDN need a
  `csp` declaration, for example
  `as_mcp_app(app, csp = list(resource_domains = "https://*.tile.openstreetmap.org"))`.
  (Leaflet's default markers don't: shinymcp puts their images in the
  page.) Fonts can't load from the app itself either, so icon fonts such
  as Font Awesome don't show; use inline SVG icons (`bsicons`, for
  example) instead.
* **Output bindings from packages**, other than htmlwidgets. Their values
  are shown as text.
* **Progress bars.** A request finishes before the page hears about its
  progress, so `withProgress()` shows nothing.
* **The page's URL.** Bookmarking, `updateQueryString()`, and
  `session$reload()` have nothing to act on.

## Sessions in a server

Sessions live in the R process that serves the app: up to 50 at a time,
each closing after an hour without use. The `shinymcp.max_views` and
`shinymcp.view_timeout` options change those limits;
`help("shinymcp-options")` lists the others.

If a request reaches a process that doesn't have its view's session (after
a restart, or behind a load balancer that spreads requests over several R
processes), the page's full input state starts a new session, and the view
carries on. What the server function kept outside its inputs, such as a
`reactiveVal()` counting clicks, starts over. `vignette("deployment")`
covers serving apps over HTTP and on Posit Connect.
