---
title: "shinymcp by example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{shinymcp by example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This article is a tour of shinymcp, one idea at a time. Each example in the package adds a single concept — inputs, outputs, native Shiny widgets, multi-tool routing, module reuse, stdio transport, and more — so you can read as far as your use case requires. Every app ships with the package and you can run any of them right now:

```{r}
shinymcp::preview_app(system.file("examples", "penguins", package = "shinymcp"))
```

## 1. hello-mcp-minimal — the shape everything builds on

Every MCP App has the same skeleton: a UI of components, a list of `ellmer::tool()` objects, an `mcp_app()` call, and `serve()`. The tool's argument names must match the input element ids; the return-list names must match the output element ids. That's the whole contract. Everything else in these examples is an elaboration of it.

```{r}
ui <- htmltools::tagList(
  mcp_text_input("name", "Your name", value = "world"),
  mcp_text("greeting")
)

tools <- list(
  ellmer::tool(
    fun = function(name = "world") {
      list(greeting = paste0("Hello, ", name, "!"))
    },
    name = "greet",
    description = "Greet a person by name",
    arguments = list(name = ellmer::type_string("The name to greet"))
  )
)

app <- mcp_app(ui, tools, name = "hello-mcp-minimal")
serve(app)
```

```{r}
shinymcp::preview_app(system.file("examples", "hello-mcp-minimal", package = "shinymcp"))
```

When not to use this pattern: a function that takes inputs and returns a scalar with no visual output is simpler as a plain MCP tool registered directly with `ellmer::chat()$register_tool()`. The `mcp_app()` layer earns its keep when you need a card the user can interact with.

## 2. hello-mcp — plot and text outputs, bslib theme

Once the basic skeleton is clear, adding richer outputs is mechanical. `mcp_plot()` expects a base64-encoded PNG in the return list; `mcp_text()` takes a string. A bslib `page()` with `bs_theme()` wraps the layout so the card inherits the host's design system.

```{r}
ui <- bslib::page(
  theme = bslib::bs_theme(preset = "shiny"),
  bslib::card(
    bslib::card_header("Dataset Explorer"),
    bslib::layout_columns(
      col_widths = c(4, 8),
      mcp_select("dataset", "Choose dataset", c("mtcars", "iris", "pressure")),
      htmltools::tagList(
        mcp_plot("plot", height = "280px"),
        mcp_text("summary")
      )
    )
  )
)
```

```{r}
shinymcp::preview_app(system.file("examples", "hello-mcp", package = "shinymcp"))
```

When not to use this pattern: if the tool returns one number or a short string, a plot card adds overhead with no benefit. Use `mcp_text()` alone until you have something worth visualizing.

## 3. penguins — native shiny/bslib inputs, auto-detected

You do not need `mcp_select()` or `mcp_checkbox()` wrappers on every input. The JS bridge reads the tool's argument names from the config and looks for matching element ids in the DOM. A standard `shiny::selectInput("species", ...)` works without any extra annotation, as long as the element id matches the tool argument name.

```{r}
sidebar = bslib::sidebar(
  width = 260,
  shiny::selectInput("species", "Species", c("All", "Adelie", "Chinstrap", "Gentoo")),
  shiny::selectInput("x_var", "X axis", var_choices),
  shiny::selectInput("y_var", "Y axis", var_choices, selected = "bill_depth_mm"),
  shiny::checkboxInput("trend", "Show trend line")
)
```

The tool arguments are `species`, `x_var`, `y_var`, and `trend`, matching the ids above exactly.

```{r}
shinymcp::preview_app(system.file("examples", "penguins", package = "shinymcp"))
```

When not to use this pattern: if you want to be explicit about which inputs are MCP-visible, or if the element id is generated by a library and doesn't match the argument name you want, the `mcp_*` wrappers in example 4 give you that control.

## 4. bslib-inputs — escape hatches for non-matching ids

The id-equals-arg-name rule covers most cases. When it doesn't — because the id is on an inner element rather than the widget container — `mcp_input()` stamps a `data-shinymcp-input` attribute on any tag. `mcp_output()` does the same for output targets.

```{r}
# radioButtons puts the id on a wrapper div; mcp_input() redirects the bridge
mcp_input(
  shiny::radioButtons("format", "Format", c("summary", "head")),
  id = "format"
)

# mcp_output() turns any tag into an output target
mcp_output(tags$pre(id = "result", style = "white-space: pre-wrap;"))
```

```{r}
shinymcp::preview_app(system.file("examples", "bslib-inputs", package = "shinymcp"))
```

When not to use this pattern: if every input in your UI already matches by id, adding `mcp_input()` wrappers is noise. Save these escape hatches for the cases where auto-detection fails.

## 5. bind-mcp-demo — annotate an existing app selectively

`bindMcp()` is a pipe-friendly annotation: add it to any Shiny input or output tag to expose that element as an MCP endpoint. Elements without `bindMcp()` continue to work in the Shiny app but are invisible to the AI agent. Then `as_mcp_app()` wraps the whole `shinyApp()` object with a tool handler.

```{r}
# This input IS exposed to MCP
selectInput("dataset", "Dataset:", c("mtcars", "iris", "pressure")) |>
  bindMcp()

# This input is NOT exposed to MCP
numericInput("n", "Rows to show:", value = 10, min = 1, max = 50)

# Wrap the existing shiny app
app <- shinyApp(ui, server) |>
  as_mcp_app(name = "dataset-explorer", tools = list(...))
serve(app)
```

```{r}
shinymcp::preview_app(system.file("examples", "bind-mcp-demo", package = "shinymcp"))
```

When not to use this pattern: if you are building from scratch, `mcp_app()` directly (as in examples 1-4) keeps the MCP contract explicit from the start. `bindMcp()` is best for incrementally adding MCP support to an app that already exists and must keep working as a plain Shiny app.

## 6. multi-tool — independent tools in one app

When a UI has two unconnected groups of inputs and outputs, they become two separate tools. The JS bridge routes each input-change event only to the tool that owns that input, rather than calling a single hardcoded function. The analyzer also traces transitive reactive chains: if `filtered_data` depends on `base_data` which depends on `input$dataset`, the tool correctly lists `dataset` as an argument even though no intermediate reactive references it directly.

```{r}
explore_data <- ellmer::tool(
  fun = function(dataset, n_rows, sort_col) { ... },
  name = "update_summary_and_plot",
  ...
)

greet <- ellmer::tool(
  fun = function(user_name) { list(greeting = ...) },
  name = "update_greeting",
  ...
)

app <- mcp_app(ui = ui, tools = list(explore_data, greet), name = "multi-tool-demo")
serve(app)
```

```{r}
shinymcp::preview_app(system.file("examples", "multi-tool", package = "shinymcp"))
```

When not to use this pattern: one task should be one tool. If you're tempted to put two conceptually unrelated operations in a single tool call, split the UI into separate `mcp_app()` objects instead.

## 7. module-tool — a Shiny module wrapped as a tool

`mcp_tool_module()` takes a standard Shiny module (a `ui` function and a `server` function) and serves it as an MCP App. The module renders using Shiny's own rendering stack; the `handler` argument provides the MCP-side function that returns the structured result the client sees.

```{r}
app <- mcp_tool_module(
  module_ui = hist_ui,
  module_server = hist_server,
  name = "histogram",
  description = "Show an interactive histogram with adjustable bins",
  handler = function(dataset = "faithful", bins = 25) {
    # ... render to PNG, return list(plot = base64_string)
    list(plot = base64enc::base64encode(tmp))
  },
  arguments = list(
    dataset = ellmer::type_string("Dataset: 'faithful' or 'mtcars'"),
    bins = ellmer::type_number("Number of histogram bins")
  )
)
serve(app)
```

```{r}
shinymcp::preview_app(system.file("examples", "module-tool", package = "shinymcp"))
```

When not to use this pattern: a one-off app has no reason to be structured as a module. Module reuse pays off when the same histogram (or form, or table) needs to appear in multiple places: a Shiny dashboard, a shinychat card, and an MCP host.

## 8. converted-dashboard — the output of `convert_app()`

This directory is what `convert_app()` produces when you point it at a plain Shiny app. The pipeline parses the source with `parse_shiny_app()`, traces the reactive graph with `analyze_reactive_graph()`, and writes files with `generate_mcp_app()`. The original app is in `original-app.R`; `app.R` is the generated result.

```{r}
# Generated output uses mcp_* components throughout
ui <- htmltools::tagList(
  mcp_select("dataset", "Dataset:", c("mtcars", "iris")),
  mcp_numeric_input("obs", "Observations:", value = 10, min = 1, max = 50),
  mcp_text("summary_text"),
  mcp_table("data_table")
)
```

```{r}
shinymcp::preview_app(system.file("examples", "converted-dashboard", package = "shinymcp"))
```

See `vignette("automatic-conversion")` for the full conversion workflow. When not to use the conversion pipeline: it infers structure but not domain logic. Review the generated tool bodies before shipping; the function bodies are mechanical translations and may need hand-tuning for complex apps.

## 9. serve-to-client — register with a real MCP client

The `serve(app, type = "stdio")` call turns the app into a stdio MCP server. You register it in your client's config and the client launches it as a subprocess. After that, the `greet` tool appears in the client's tool list and calling it renders the card inline.

```{r}
app <- mcp_app(ui = ..., tools = list(...), name = "shinymcp-demo")
serve(app, type = "stdio")
```

To connect this to Claude Desktop or VS Code, find the absolute path to `serve.R` and follow the instructions in `inst/examples/serve-to-client/README.md`. The key step is adding a `mcpServers` entry that points to this file:

```json
{
  "mcpServers": {
    "shinymcp-demo": {
      "command": "Rscript",
      "args": ["/ABSOLUTE/PATH/TO/serve-to-client/serve.R"]
    }
  }
}
```

When not to use this pattern: for local iteration and UI development, `preview_app()` is faster. Switch to stdio once the tool contract is settled and you want to test the end-to-end client experience.

## 10. shinychat-card — surface a card in shinychat

`as_shinychat_tool()` wraps an `mcp_app()` object as a shinychat tool result. When an LLM calls the tool, shinychat renders the card inline in the chat UI. `mcp_result_plot()` packages a base64 PNG with an alt-text string so the result is portable to non-visual clients.

```{r}
card_app <- mcp_app(
  ui = htmltools::tagList(
    mcp_select("dataset", "Dataset", c("mtcars", "iris", "pressure")),
    mcp_text("summary"),
    mcp_plot("plot", height = "220px")
  ),
  tools = list(list(
    name = "inspect_dataset",
    fun = function(dataset = "mtcars") {
      # summary_text <- paste(capture.output(summary(get(dataset))), collapse = "\n")
      list(
        summary = summary_text,
        plot = mcp_result_plot(function() { ... }, text = paste("Scatter plot for", dataset))
      )
    },
    ...
  )),
  name = "shinychat-card-demo"
)

inspect_dataset <- as_shinychat_tool(
  card_app,
  value_fn = function(raw_result) list(summary = raw_result$summary),
  summary = function(raw_result) raw_result$summary,
  title = "Dataset Inspector"
)
```

```{r}
shiny::runApp(system.file("examples", "shinychat-card", package = "shinymcp"))
```

When not to use this pattern: if you are not building a chat UI with shinychat, you do not need `as_shinychat_tool()`. The plain MCP tool registered with `serve()` or `mcp_host_server()` is enough.

## 11. embed-in-shiny — host a card in a dashboard

`mcp_host_ui()` and `mcp_host_server()` let you embed any `mcp_app()` object directly inside a Shiny dashboard as a sandboxed iframe. Define the app once, then use it either as a client-facing MCP server or as an embedded review panel; the same object works in both contexts.

```{r}
greet_app <- mcp_app(ui = ..., tools = list(...), name = "greet")

ui <- bslib::page_sidebar(
  title = "An MCP App embedded in Shiny",
  sidebar = bslib::sidebar(...),
  bslib::card(
    bslib::card_header("Embedded MCP App"),
    mcp_host_ui("greet")
  )
)

server <- function(input, output, session) {
  mcp_host_server("greet", greet_app, trigger = "change")
}

shinyApp(ui, server)
```

```{r}
shiny::runApp(system.file("examples", "embed-in-shiny", package = "shinymcp"))
```

When not to use this pattern: if you are deploying directly to an MCP client with no Shiny host involved, `serve(app, type = "stdio")` (example 9) is all you need. The Shiny host is for review surfaces, dashboards, and hybrid applications where a human oversees what the agent produced.

## 12. data-explorer — build inputs from a data frame

When the schema of a data frame drives the UI, compute the input choices from the frame itself rather than hard-coding them. The auto-detection rule from example 3 still applies: the column names you use as select choices become the argument values, and the argument names match the element ids.

```{r}
numeric_cols <- names(data)[vapply(data, is.numeric, logical(1))]
categorical_cols <- names(data)[vapply(
  data,
  function(x) is.character(x) || is.factor(x) || length(unique(x)) <= 10,
  logical(1)
)]

# Build choices programmatically
num_choices <- setNames(numeric_cols, gsub("_", " ", numeric_cols))

# Feed them into standard shiny inputs
shiny::selectInput("x_var", "X axis", num_choices)
shiny::selectInput("y_var", "Y axis", num_choices, selected = numeric_cols[2])
```

```{r}
shinymcp::preview_app(system.file("examples", "data-explorer", package = "shinymcp"))
```

When not to use this pattern: a fixed, well-understood schema is clearer hard-coded. Programmatic inputs are worth the extra code only when the columns genuinely change between deployments or users.

## 13. rpharma-hangout — capstone: multi-skill clinical app

The R/Pharma demo puts everything together. Two `mcp_app()` objects (a Safety Signal Scout and an Enrollment Rescue Simulator) run as selectable skill cards inside a clinical Shiny dashboard. Each produces a `mcp_result_text()` return with a `model_value` list; the parent Shiny app reads that structured data from `mcp_host_server()$last_raw_result()` without scraping the display. A contract inspector panel renders the tool's input schema live, and both cards are also registered with shinychat via `as_shinychat_tool()`.

```{r}
# Each skill card carries a structured model_value alongside display content
list(
  memo = mcp_result_text(
    memo,
    model_value = list(
      widget = "Safety Signal Scout",
      decision = decision,
      risk_ratio = rr,
      ...
    )
  ),
  evidence = mcp_result_table(evidence, text = "Aggregate AE evidence table."),
  risk_plot = mcp_result_plot(function() { ... }, model_value = list(...), text = "...")
)
```

```{r}
shiny::runApp(system.file("examples", "rpharma-hangout", package = "shinymcp"))
```

See `inst/examples/rpharma-hangout/README.md` for the full setup. The clinical data is synthetic; the patterns (approved skills, inspectable contracts, structured handoffs) apply to any regulated or high-stakes domain where a human needs to see exactly what the agent called and what it got back.

When not to use this pattern: if you do not need human oversight of agent calls and the app covers a single skill, a single `mcp_app()` served directly (example 9) is far less setup than a multi-card dashboard with a shinychat integration.

## Going further

These examples keep the LLM out of the loop so they run offline. Two related packages build
on the same R functions:

- [`{dsprrr}`](https://jameshwade.github.io/dsprrr/) fits prompts to data with DSPy-style
  signatures, optimization, and tracing, so you improve a prompt with examples instead of by hand.
- [`{deputy}`](https://jameshwade.github.io/deputy/) is an agent runtime: it lets a tool-using
  agent call reviewed R functions, with permissions, hooks, and multi-agent coordination.
