---
title: "Building an app from tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building an app from tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("ellmer", quietly = TRUE)
)
```

An app made with `mcp_app()` has two halves: tools, which are R functions
the model can call on its own, and a page of inputs and outputs that calls
the same functions when the person using it changes something. Nothing
runs between calls. Every answer depends only on its arguments, so any R
process can give it, and the tools are as useful in chat clients that
can't show the page, or to other agents, as in ones that can.

To start from a Shiny app you already have, see
`vignette("rewriting-as-tools")`.

## A first app

```{r first-app, message = FALSE}
library(shinymcp)

summarize_dataset <- ellmer::tool(
  function(dataset = "mtcars") {
    data <- getExportedValue("datasets", dataset)
    list(
      rows = nrow(data),
      summary = paste(capture.output(summary(data)), collapse = "\n")
    )
  },
  name = "summarize_dataset",
  description = "Summarize one of R's built-in datasets, column by column.",
  arguments = list(
    dataset = ellmer::type_enum(c("mtcars", "faithful"), "The dataset to summarize.")
  )
)

app <- mcp_app(
  ui = htmltools::tagList(
    mcp_select("dataset", "Dataset", c("mtcars", "faithful")),
    mcp_text("rows"),
    mcp_text("summary")
  ),
  tools = list(summarize_dataset),
  name = "datasets"
)
```

Names connect the two halves. The tool's argument `dataset` takes its value
from the input with id `dataset`, and the list the tool returns fills the
outputs with ids `rows` and `summary`. When the person picks another
dataset, the page calls every tool that takes `dataset` and fills in what
they return.

`$call_tool()` runs a tool and returns its R value, which is what you want
in tests:

```{r call-tool}
str(app$call_tool("summarize_dataset", list(dataset = "faithful")))
```

`$run_tool()` returns what an MCP client receives:

```{r run-tool}
result <- app$run_tool("summarize_dataset", list(dataset = "faithful"))
cat(result$content[[1]]$text)
```

Try it with `preview_app(app)`, and see `vignette("deployment")` for putting
it in a chat client.

## Inputs

Any input that has the argument's name as its id works: Shiny's
(`selectInput()`, `sliderInput()`, `dateRangeInput()`, ...), bslib's, and
the small ones shinymcp provides, `mcp_select()` and its relatives, which
need neither package. When an id can't match the argument, because two
tools share the page or a module prefixed it, mark the element with
`mcp_input(tag, id = "argument")`.

By default the page calls tools a quarter of a second after the person
stops changing inputs. `mcp_app(trigger = "change")` calls on every change,
and `trigger = "submit"` waits for an apply button, which
`mcp_submit_button()` places (otherwise the page adds one at the bottom).
Buttons made with `mcp_action_button()` or `shiny::actionButton()` call the
tools that take their id when pressed. Those tools run only then, as an
`eventReactive()` would, and not when their other inputs change.

## Outputs and results

`mcp_text()`, `mcp_plot()`, `mcp_table()`, and `mcp_html()` make the
elements results are drawn into. Shiny's output functions work too, matched
by id the same way. A plot is drawn at the size of its output: the page
tells each tool call how big its plot outputs are, and calls again when
one changes size. A call the page didn't make, such as the model's, draws
at 800 by 500 pixels, which the page shows scaled to fit until it has asked
for the right size. Give `mcp_result_plot()` a `width` and `height` to
draw at a size of your own.

A tool can return plain values: a string, a number, a data frame, an
htmltools tag, a ggplot object. The `mcp_result_*()` functions say more
about a value and what the model should read for it:

```{r typed, eval = FALSE}
list(
  summary = mcp_result_text(text, model_value = list(n = nrow(data))),
  table = mcp_result_table(head(data, 20)),
  plot = mcp_result_plot(function() hist(data$mpg), text = "Histogram of mpg"),
  report = mcp_result_pdf("report.pdf")
)
```

Each result is split three ways, because it has three readers:

* **Text** in `content`, for the model and for clients that can't show the
  app. By default it joins each output's text.
* **Structured data** in `structuredContent`, for the model: one value per
  output, such as a table's rows or the `model_value` you gave.
* **What the page draws** (HTML, images, widget data and their JavaScript)
  in `_meta`, which clients pass to the page and keep from the model.

`mcp_tool_result()` writes the first two yourself, for example to lead with
the identifiers the model should quote back, or to report an error the
model should act on:

```{r tool-result, eval = FALSE}
mcp_tool_result(
  answer = sprintf("%d per group.", n),
  curve = mcp_result_plot(draw_curve),
  text = sprintf("The study needs %d participants per group.", n),
  data = list(n_per_group = n, power = power)
)
```

The [`sample-size`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/sample-size)
example is a complete app in this style.

## Tools

`ellmer::tool()` is the most direct way to write a tool: its argument types
become the JSON Schema the model sees, and its annotations
(`read_only_hint` and the rest) tell clients what the tool does. Plain lists
work too, with `name`, `description`, `fun`, and optionally `inputSchema`,
`outputSchema`, and `annotations`.

The model's arguments are checked against the schema before the function
runs: required arguments must be there, and values must have the declared
type and, for choices, one of the listed values. A problem goes back to the
model as a tool error it can correct. Arguments you pass from R, to
`$call_tool()` for instance, are read as a client would send them: a vector
of one value is one value, not an array, so write an array of one as
`list(x)`.

Some tools shouldn't be the model's to call: saving a result, sending an
order, anything the person should decide. `tool_visibility` hides a tool
from the model while the page can still call it. Give such a tool the id
of a button as an argument, and it runs when the person presses the
button, with the other inputs it takes as they are:

```{r visibility, eval = FALSE}
save_choice <- ellmer::tool(
  function(save, dataset) {
    saveRDS(dataset, "choice.rds")
    list(status = paste("Saved", dataset))
  },
  name = "save_choice",
  description = "Save the chosen dataset.",
  arguments = list(
    save = ellmer::type_integer("Times the Save button was pressed."),
    dataset = ellmer::type_string("The dataset to save.")
  ),
  annotations = ellmer::tool_annotations(read_only_hint = FALSE)
)

mcp_app(
  ui = htmltools::tagList(
    mcp_select("dataset", "Dataset", c("mtcars", "faithful")),
    mcp_action_button("save", "Save"),
    mcp_text("status")
  ),
  tools = list(save_choice),
  tool_visibility = list(save_choice = "app")
)
```

The page never runs a tool annotated with `read_only_hint = FALSE` or
`destructive_hint = TRUE` on its own, to fill in outputs when it opens.

`tool_outputs` declares the outputs each tool returns
(`list(explore = c("scatter", "stats"))`), which gives it an `outputSchema`.

## Telling the model what the person did

After the person changes something, the page tells the model on its next
turn which values they chose. Set `mcp_app(model_context = FALSE)` to turn
that off. The page's JavaScript can also do it directly, with
`window.shinymcp.updateModelContext()`, and post a message into the chat
with `window.shinymcp.sendMessage()`.

## Look and feel

The UI can be a whole page, such as `bslib::page_sidebar()`, or a fragment;
give a fragment a theme with `mcp_app(theme = bslib::bs_theme())`. By
default the page takes the chat client's colors and fonts when it provides
them, so the app matches the conversation around it; `host_styles = FALSE`
keeps your own. Pages built on Bootstrap 5 (bslib) also follow the client's
dark mode.

Everything the page needs is written into it, because chat clients show
apps under a Content Security Policy that blocks other sources. Files the
UI loads by relative path come from `www`
(`mcp_app(ui, tools, www = "www")`). Anything from another site needs a
`csp` declaration, such as
`csp = list(resource_domains = "https://cdn.example.com")`.

## Loading data on demand

A page that needs more data than a result should carry (a large table, a
map layer) can read it when it needs it. Declare it with `resources`, and
read it from the page's JavaScript with `window.shinymcp.readResource()`:

```{r resources, eval = FALSE}
mcp_app(
  ui,
  tools,
  resources = list(
    "ui://datasets/stations.json" = function() jsonlite::toJSON(stations)
  )
)
```

The [`feature-tour`](https://github.com/JamesHWade/shinymcp/tree/main/inst/examples/feature-tour)
example uses this and the rest of the page's JavaScript API.
