---
title: "Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Most problems are easier to see in `preview_app()` than in a chat client.
It shows the app as clients do, under the same Content Security Policy, and
next to it what the model received, every tool call with its result, the
protocol messages, and the server's tools. The browser's developer console
shows the page's own errors; the preview's page is in an iframe, so choose
it in the console's context menu.

In R, `app$run_tool("name", list(...))` returns exactly what a client would
receive, and `app$call_tool()` the tool's R value.

## The client doesn't list the app's tools

The client couldn't start the server, or the server stopped.

* Run the command the client runs, `Rscript /full/path/to/app.R`, in a
  terminal. It should wait for input without printing anything (Ctrl+C
  stops it). An error here is the one the client hit.
* Use full paths in the client's settings. Desktop clients don't start in
  your home directory, and on macOS they don't see your shell's `PATH`, so
  `Rscript` may need its full path too (`which Rscript`).
* The packages the app uses must be installed in the library that
  `Rscript` uses. With renv, point the client at the project's library or
  activate it in the script.
* Nothing may print to standard output before `serve()` starts: the
  protocol runs over it. Messages and warnings go to standard error and are
  fine.
* Check the client's log. Claude Desktop keeps one for each server, on
  macOS in `~/Library/Logs/Claude/`.

## The tools work but the app doesn't show

The client doesn't support MCP Apps, or not in this context. Terminal
clients such as Claude Code show text only. shinymcp gives those clients
the same tools, with text results.

## Something on the page is missing or blank

Chat clients block anything the app doesn't carry itself.

* Content from another site (map tiles, a script from a CDN, a web font)
  needs a `csp` declaration: `csp = list(resource_domains = "https://tiles.example.com")`
  for images, scripts, and stylesheets, `connect_domains` for `fetch()`.
  The preview's console lists each blocked request.
* Files the UI loads by relative path must be in the app's `www/` folder
  or under a path added with `shiny::addResourcePath()`.
* Fonts don't load from the app itself, so icon fonts such as Font
  Awesome show nothing. Use SVG icons (the bsicons package, for example).

## An input doesn't reach the tool

* A tool argument takes the value of the input with the same id.
  `mcp_input(tag, id = "argument")` connects an input whose id differs.
* An input from a package works if the package registers a Shiny input
  binding for it. If it uses JavaScript without one, call
  `Shiny.setInputValue("id", value)` from that JavaScript.
* Inputs in HTML that a tool returns aren't connected to tools. Put them
  in the UI from the start.
* In a live Shiny app with anything marked by `bindMcp()`, the model can
  only set marked inputs. The person can still change all of them.

## A conditional panel doesn't show or hide

shinymcp reads a `conditionalPanel()`'s condition itself, since chat
clients forbid running it as JavaScript. A condition it can't read shows
its panel and logs a warning in the browser's console. Conditions built
from comparisons, `&&`, `||`, `!`, `input.x`, `output.x`, and methods such
as `indexOf()` work; functions defined elsewhere on the page don't.

## An output doesn't update

* The names of the list a tool returns must match output ids.
  `mcp_output(tag, id = "name")` connects an output whose id differs.
* In a live Shiny app, `req()` that fails clears an output, as in Shiny;
  `validate()` shows its message in it. An error in a render function
  shows in the output, and in the text the model receives.

## A live Shiny app stops with an error

An error in an observer ends a Shiny session, in a browser too. shinymcp
reports it: a tool call from the model returns it as an error, and the page
shows it above the app. The R console (or the client's log, for a stdio
server) has the stack trace. The page's next change starts a new session.

## A live Shiny app forgets what the person did

Each view keeps its session in the R process that opened it, for up to an
hour without use, and a process keeps at most 50 views (the
`shinymcp.view_timeout` and `shinymcp.max_views` options). When a view's
session is gone, the page starts a new one from its inputs: what the server
function kept outside its inputs starts over. On Posit Connect, this also
happens when requests reach another process; set **Max processes** to 1 for
apps that keep such state.

## An upload is refused

Uploads work in live Shiny apps; a tool can't take a file from the page.
They are limited by Shiny's `shiny.maxRequestSize` option, 5 MB unless
the app sets it higher:

```{r}
options(shiny.maxRequestSize = 30 * 1024^2)
```

## The model doesn't know what the person did

The page tells the model through the client, which must support
`ui/update-model-context`. The preview's Context tab shows what the app
sent. `mcp_app(model_context = FALSE)` turns it off; in a live Shiny app,
`mcp_model_context()` replaces the default summary.

In a shinychat conversation, `mcp_chat_host()` passes it on before each
message the person sends, as a message of its own; `as_shinychat_tool()`
alone doesn't. Some providers, such as AWS Bedrock, refuse two user
messages in a row. For those, set `context = FALSE` and put `host$context()`
in the person's message yourself.

## An app in a Shiny app doesn't show

A card or pane that can't show its app says why, above where the app would
be:

* *Couldn't reach the app's server*, or *Couldn't load the app*: the URL
  given to `mcp_client()` is wrong, the server isn't running, or it wants
  credentials in `headers`. `client$tools()` in the console fails with the
  same error, in more detail.
* *This app isn't available any more*: a card restored with a saved
  conversation, whose server this session doesn't host, or which no longer
  has the card's tool. Give `mcp_chat_host()` the same sources in every
  session, with the same names. A client is named after its URL unless you
  give it a `name`.
* *... has no tool called ...*, or *The tool ... doesn't show an app*: a
  pane names a tool its server doesn't have, or one that declares no page.
  For an app in the same R process, `mcp_host_server()` stops with this
  error instead.

An app that holds requests open while it waits for changes, as a Shiny app
served with Shiny's own MCP support can, needs a client `timeout` longer
than that wait.
