---
title: "Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Most problems are easier to see in `preview_app()` than in a chat client.
It shows the app as clients do, under the same Content Security Policy, and
next to it what the model received, every tool call with its result, the
protocol messages, and the server's tools. The browser's developer console
shows the page's own errors; the preview's page is in an iframe, so choose
it in the console's context menu.

In R, `app$run_tool("name", list(...))` returns exactly what a client would
receive, and `app$call_tool()` the tool's R value.

## The client doesn't list the app's tools

The client couldn't start the server, or the server stopped.

* Run the command the client runs, `Rscript /full/path/to/app.R`, in a
  terminal. It should wait for input without printing anything (Ctrl+C
  stops it). An error here is the one the client hit.
* Use full paths in the client's settings. Desktop clients don't start in
  your home directory, and on macOS they don't see your shell's `PATH`, so
  `Rscript` may need its full path too (`which Rscript`).
* The packages the app uses must be installed in the library that
  `Rscript` uses. With renv, point the client at the project's library or
  activate it in the script.
* Nothing may print to standard output before `serve()` starts: the
  protocol runs over it. Messages and warnings go to standard error and are
  fine.
* Check the client's log. Claude Desktop keeps one for each server, on
  macOS in `~/Library/Logs/Claude/`.

## The tools work but the app doesn't show

The client doesn't support MCP Apps, or not in this context. Terminal
clients such as Claude Code show text only. shinymcp gives those clients
the same tools, with text results.

## Something on the page is missing or blank

Chat clients block anything the app doesn't carry itself.

* Content from another site (map tiles, a script from a CDN, a web font)
  needs a `csp` declaration: `csp = list(resource_domains = "https://tiles.example.com")`
  for images, scripts, and stylesheets, `connect_domains` for `fetch()`.
  The preview's console lists each blocked request.
* Files the UI loads by relative path must be in the app's `www/` folder
  or under a path added with `shiny::addResourcePath()`.
* Fonts don't load from the app itself, so icon fonts such as Font
  Awesome show nothing. Use SVG icons (the bsicons package, for example).

## An input doesn't reach the server

* In a Shiny app with anything marked by `bindMcp()`, the model can only
  set marked inputs. The person can still change all of them.
* An input from a package works if the package registers a Shiny input
  binding for it. If it uses JavaScript without one, call
  `Shiny.setInputValue("id", value)` from that JavaScript.
* In an app built from tools, a tool argument takes the value of the input
  with the same id. `mcp_input(tag, id = "argument")` connects an input
  whose id differs.

## An output doesn't update

* In an app built from tools, the names of the list a tool returns must
  match output ids.
* `req()` that fails clears an output, as in Shiny; `validate()` shows its
  message in it.
* An error in a render function shows in the output, and in the text the
  model receives.

## The app stops with an error

An error in an observer ends a Shiny session, in a browser too. shinymcp
reports it: a tool call from the model returns it as an error, and the page
shows it above the app. The R console (or the client's log, for a stdio
server) has the stack trace. The page's next change starts a new session.

## The app forgets what the person did

Each view keeps its session in the R process that opened it, for up to an
hour without use, and a process keeps at most 50 views (the
`shinymcp.view_timeout` and `shinymcp.max_views` options). When a view's
session is gone, the page starts a new one from its inputs: what the server
function kept outside its inputs starts over. On Posit Connect, this also
happens when requests reach another process; set **Max processes** to 1 for
apps that keep such state.

## An upload is refused

Uploads are limited by Shiny's `shiny.maxRequestSize` option, 5 MB unless
the app sets it higher:

```{r}
options(shiny.maxRequestSize = 30 * 1024^2)
```

## The model doesn't know what the person did

The page tells the model through the client, which must support
`ui/update-model-context`. The preview's Context tab shows what the app
sent. In a Shiny app, `mcp_model_context()` replaces the default summary;
`mcp_app(model_context = FALSE)` turns it off.
