---
title: "shinymcp use cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{shinymcp use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`shinymcp` is strongest when an AI assistant needs to do more than print a
number. The assistant can call a tool, receive structured data for reasoning,
and show the user an interactive card with controls, tables, plots, and
formatted output — and because the card reports the user's interactions back
into the model's context, the assistant can act on what the user did in its
next turn.

## The example ladder

The bundled examples are ordered so each one adds a single new idea. Work
through them in order and you'll have touched every major feature of the
package. Each lives at
`system.file("examples", "<name>", "app.R", package = "shinymcp")`, and each
runs in `preview_app()` — a local reference host with a protocol log — or as
a stdio MCP server via `Rscript app.R`.

| Level | Example | The one new idea |
|---|---|---|
| 1 | `hello-mcp` | The whole contract: one input, one tool, one output, ~30 lines |
| 2 | `bslib-inputs` | Native shiny/bslib inputs auto-detected by id — no wrappers |
| 3 | `penguins` | A real dashboard: ggplot2, multiple outputs, a declared `outputSchema` |
| 4 | `feature-tour` | The protocol features: app-only tools, lazy resources, theme syncing, `window.shinymcp` |
| 5 | `multi-tool` | One app, several tools — connected reactive groups become separate contracts |
| 6 | `use-cases` gallery | Realistic chat cards with typed `model_value` handoffs (this article) |
| 7 | `shinychat-card` + `rpharma-hangout` | Embedding: the same `McpApp` inside Shiny dashboards and shinychat |

Three ideas carry the whole ladder:

1. **Names are the contract.** Tool argument names match input ids; result
   keys match output ids. That's the entire binding model.
2. **One result, two consumers.** The human gets the rendered card; the model
   gets structured values it can reason over without scraping the display.
3. **One definition, many homes.** The same `McpApp` serves to Claude over
   MCP, embeds in a Shiny app, and wraps as a shinychat tool card.

## Level 4 in detail: the feature tour

`feature-tour` packs the MCP Apps protocol features into one annotated app.
Open it in `preview_app()` with the protocol log visible and you can watch
each one on the wire:

```{r}
source(system.file("examples", "feature-tour", "app.R", package = "shinymcp"))
```

What to look for:

- **App-only tools** (`tool_visibility`): the `fetch_region_detail` tool is
  callable from the card's "Load region detail" button but carries
  `_meta.ui.visibility = ["app"]`, so the model never sees it. Use this to
  keep UI plumbing — or human-only detail views — out of the model's tool
  list.
- **Lazy resources** (`resources` + `window.shinymcp.readResource()`): the
  region catalog isn't inlined into the HTML; the card fetches it through
  the host at startup. This is the pattern for shipping large datasets
  without bloating the `ui://` resource.
- **Declared result shapes** (`tool_outputs`): the analysis tool publishes
  an `outputSchema`, so hosts and models know it returns a base64 plot named
  `trend` and a text summary named `summary` before ever calling it.
- **Theme syncing**: flip your OS or host theme and the card follows —
  the bridge maps the host's theme onto Bootstrap's `data-bs-theme`.
- **Host interactions** (`window.shinymcp`): buttons that send a message
  into the conversation, open a link, and request fullscreen — all spec
  methods, all degrading gracefully when a host doesn't support them.

## The use-case gallery

The gallery contains three small but realistic MCP Apps:

| Use case | What it demonstrates |
|---|---|
| Revenue scenario board | Scenario controls, model-facing structured values, a table, and an ARR plot |
| Experiment planner | Statistical planning, a power curve, and a concise recommendation |
| Incident triage console | Operational decision support, status HTML, and a runbook table |

The apps are intentionally card-sized: controls fit in a compact grid, the
primary answer is immediately visible, and secondary tables stay available
without taking over the chat transcript. Each declares `tool_outputs` (so
its result shape is part of the published contract) and `prefers_border`
(a rendering hint for the host).

### Run a use case as an MCP App

The gallery entrypoint serves one use case at a time. Set
`SHINYMCP_USE_CASE` to `revenue`, `experiment`, or `incident`.

```{bash}
SHINYMCP_USE_CASE=revenue Rscript \
  "$(Rscript -e 'cat(system.file("examples", "use-cases", "app.R", package = "shinymcp"))')"
```

For Claude Desktop or another MCP host, point the server command at that same
`app.R` file and set the environment variable for the use case you want to
show.

### Revenue scenario board

This card turns go-to-market assumptions into a twelve-month forecast. The
assistant can reason over the structured result while the user sees the ARR
trajectory and monthly forecast table.

```{r}
source(system.file("examples", "use-cases", "apps.R", package = "shinymcp"))
app <- shinymcp_use_case("revenue")
app$call_tool("forecast_revenue", list(
  segment = "Mid-market",
  visitors = 25000,
  trial_rate = 7,
  win_rate = 22,
  contract_value = 9000,
  monthly_churn = 2.5
))
```

### Experiment planner

This card helps an assistant move from "can we detect a 15% lift?" to a
specific sample size and runtime. The plot gives the human reviewer a quick
sense of how much margin the design has.

```{r}
source(system.file("examples", "use-cases", "apps.R", package = "shinymcp"))
app <- shinymcp_use_case("experiment")
app$call_tool("plan_experiment", list(
  baseline_rate = 12,
  minimum_effect = 15,
  target_power = 0.9,
  traffic_per_day = 6000
))
```

### Incident triage console

This card is useful for support, SRE, and operations chats. The assistant can
collect facts conversationally, call the tool, and return a priority, briefing,
and runbook without losing the structured values it needs for follow-up.

```{r}
source(system.file("examples", "use-cases", "apps.R", package = "shinymcp"))
app <- shinymcp_use_case("incident")
app$call_tool("triage_incident", list(
  service = "Payments",
  severity = "Degraded",
  affected_users = 1200,
  minutes_open = 24,
  regulated_data = TRUE
))
```

## shinychat integration

The gallery also includes a Shiny app that registers all three cards as
`shinychat` tool results through `as_shinychat_tool()`.

```{bash}
Rscript "$(Rscript -e 'cat(system.file("examples", "use-cases", "shinychat-app.R", package = "shinymcp"))')"
```

When `OPENAI_API_KEY` is available, that app uses `chat_mod_ui()` and
`chat_mod_server()` with one ellmer client per Shiny session. Without a
configured model provider, it falls back to local demo mode: type `revenue`,
`experiment`, or `incident` and the app appends a live shinymcp card directly.

The Shiny session is still the runtime for those cards: it registers the
`McpApp`, receives bridge events from the iframe, and executes tool calls in R.
The iframe itself remains a portable MCP App surface instead of a nested Shiny
app, which keeps each chat card small and reusable outside Shiny.

For the full embedding-and-governance story — contract inspectors, typed
meeting-note handoffs, approved skill registries — see the R/Pharma demo at
`system.file("examples", "rpharma-hangout", package = "shinymcp")`.

## Why these examples matter

These examples cover the surfaces most teams need before adopting MCP Apps:

- Rich UI for people: inputs, tables, plots, and formatted HTML.
- Structured values for the model: the assistant can reason over numbers
  without scraping the display — and `ui/update-model-context` keeps it
  aware of what the user changed in the card.
- Declared contracts: input schemas, output schemas, visibility scoping,
  and resource metadata that a reviewer can inspect before anything runs.
- Portable deployment: the same `McpApp` can be served to an MCP host,
  embedded in Shiny, or wrapped as a shinychat tool card — and in hosts
  without MCP Apps support, the tools still work as plain text tools.
- Small cards over giant dashboards: each card has one job and maps cleanly
  to one tool call.
