---
title: "Use shinymcp with shinychat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use shinymcp with shinychat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`shinymcp` supports two runtime-first shinychat paths:

1. Wrap a small `McpApp` as an ellmer tool with `as_shinychat_tool()`.
2. Embed a live `McpApp` directly in Shiny with `mcp_host_ui()` / `mcp_host_server()` or `mcp_embed()`.

In both paths, the card is a portable MCP App iframe. When it runs inside
Shiny, the active Shiny session owns the live host state and executes tool
calls in R. The iframe stays lightweight HTML plus the shinymcp bridge rather
than starting a nested Shiny runtime for every chat card.

`shinymcp` also sets shinychat's `display$full_screen` flag for tool cards when
available, and the embedded host shell includes its own full-screen control for
direct Shiny embeds and older shinychat development builds.

## Tool-card path with `chat_mod_server()`

```{r}
library(shiny)
library(bslib)
library(ellmer)
library(shinychat)
library(shinymcp)

card_app <- mcp_app(
  ui = htmltools::tagList(
    mcp_text_input("name", "Name"),
    mcp_text("greeting")
  ),
  tools = list(
    list(
      name = "greet",
      description = "Generate a greeting card",
      inputSchema = list(
        type = "object",
        properties = list(name = list(type = "string"))
      ),
      fun = function(name = "world") {
        list(greeting = paste("Hello", name))
      }
    )
  ),
  name = "greeting-card"
)

greet_tool <- as_shinychat_tool(
  card_app,
  value_fn = function(raw_result) list(greeting = raw_result$greeting),
  summary = function(raw_result) raw_result$greeting,
  title = "Greeting Card"
)

ui <- page_fillable(chat_mod_ui("chat"))

server <- function(input, output, session) {
  client <- ellmer::chat("openai/gpt-4.1-nano")
  client$register_tool(greet_tool)
  chat_mod_server("chat", client)
}

shinyApp(ui, server)
```

The wrapped tool returns:

- A machine-facing value, which defaults to the raw tool result transformed by `value_fn`.
- A human-facing shinychat tool card, which defaults to a live embedded `McpApp`.
- Full-screen affordances for larger card inspection, using shinychat's native tool-card mode when available.

## Content-streaming path with `chat_ui()` + `chat_append()`

Use `mcp_content_result()` when you already have a live Shiny session and want to append a card yourself:

```{r}
ui <- page_fillable(chat_ui("chat"))

server <- function(input, output, session) {
  observeEvent(input$chat_user_input, {
    card <- mcp_content_result(
      app = card_app,
      value = list(status = "ready"),
      title = "Greeting Card"
    )

    chat_append("chat", card)
  })
}
```

## Direct embedding in Shiny

For authored Shiny UIs, use the host shell helpers directly:

```{r}
ui <- page_fillable(
  mcp_host_ui("card")
)

server <- function(input, output, session) {
  host <- mcp_host_server(
    "card",
    app = card_app,
    trigger = "submit"
  )
}
```

`mcp_embed(card_app)` is the convenience helper for dynamic server-side contexts where a live session already exists.

The returned host object also exposes read-only reactives so the outer Shiny app
can observe the embedded card:

```{r}
server <- function(input, output, session) {
  host <- mcp_host_server("card", app = card_app)

  observeEvent(host$model_context(), {
    # Current values the card is sending to the model context.
    str(host$model_context())
  })

  observeEvent(host$last_tool_call(), {
    # Raw R result plus formatted MCP structuredContent.
    str(host$last_raw_result())
    str(host$last_result()$structuredContent)
  })
}
```

## Interaction modes

`mcp_embed()` and `mcp_host_server()` support four trigger modes:

- `change`: execute immediately on input change.
- `debounce`: execute after a short quiet period.
- `submit`: require the host shell Apply button.
- `manual`: require an explicit host-side command such as `host$execute()`.

## Recommended usage

- Prefer small single-purpose cards.
- Prefer sequential tool execution in shinychat until you have a strong reason to expose many parallel cards.
- Treat the chat card as the human surface and the `value_fn` output as the model-facing value.
- Use Shiny modules directly when you need a Shiny-only app with full reactive UI semantics.
- Use `McpApp` cards when you want the same experience to work in shinychat and MCP hosts.
